### 2023-04-26 manuscript/paper.Rmd
Manuscript skeleton.

---
title: |
  |  \Large Optimal K: some titles
author: |
 | \large Mena Whalen 
  \vspace{-1.1mm}
   \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
   | \normalsize Chicago, IL 60660 \vspace{-1mm}
   | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}

   \large Gregory J. Matthews 
  \vspace{-1.1mm}
   \normalsize Department of Mathematics and Statistics \vspace{-1mm}
   \normalsize Loyola University Chicago \vspace{-1mm}
   \normalsize Chicago, IL 60660 \vspace{-1mm}
   \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}

abstract: |
  | Abstract \vspace{2mm}
  | *Keywords*: Cross validation  
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

\newpage

# Introduction {#sec:intro}
We wanted to look at how the choice of k in k-fold cross validation affected the estimation of the prediction error.

Literature review.

# Simulation Study
This is how you cite a paper: @R2022language

Simulations were performed using the R programming language (@R2022language).

Describe simulation studty with LASSO.

# Results
We find blah blah

#Conclusions 
Early work found this.  More recent work found this.  Our work confirms that.  We suggest........

Future work?  What else would we like to do?

# Acknowledgements {-}

Everyone who took STAT338 in Fall 2022.

# Supplementary Material {-}

All code for reproducing the analyses in this paper is publicly available at https://github.com/gjm112/optimal_k

# References

### 2023-05-06 simulations/simulation_lasso_parallel.R
“testing”

@@ -25,9 +25,13 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- 1000
+n <- 750 # 250 # 500 # 100 # 1000
 nsim <- 1000
-k <- seq(2, 100, 2) # c(2, 10, 100, 500, 1000)
+# k <- seq(2, 100, 2) # for n = 100 and n = 1000
+# k <- c(2, seq(10, 250, 10)) # for n = 250
+k <- c(2, seq(10, 500, 10)) # for n = 500
+# k <- c(2, seq(10, 750, 10)) # for n = 750
+# k <- c(2, 10, 100, 500, 1000) # for n = 1000
 n_mod <- length(variables)
 
 # create objects
@@ -66,7 +70,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_100.rds")
+out <- readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_500_n_500.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -78,7 +82,7 @@ mse <- apply((theta.hat - theta)^2, c(1, 2), mean)
 library(plotrix)
 
 # bias^2 plot
-plot(k, biassq[, 1], type = "b", xlim = c(0, 20), ylim = c(0, 35), ylab = "bias^2")
+plot(k, biassq[, 1], type = "b", ylim = c(min(biassq), max(biassq)), ylab = "bias^2")
 points(k, biassq[, 2], type = "b", col = "blue")
 points(k, biassq[, 3], type = "b", col = "red")
 points(k, biassq[, 4], type = "b", col = "purple")
@@ -87,7 +91,7 @@ legend(800, 20, legend=c("1:3", "1:5", "1:20", "1:100", "6:100"),
        col=c("black", "blue", "red", "purple", "green"), lty = 1, cex = 0.8)
 
 # variance plot
-plot(k, var[, 1], type = "b", xlim = c(0, 20), ylim = c(20, 23), ylab = "var")
+plot(k, var[, 1], type = "b", ylim = c(min(var), max(var)), ylab = "var")
 points(k, var[, 2], type = "b", col = "blue")
[…]

### 2023-05-06 manuscript/paper.Rmd
check,

@@
 ---
-title: |
-  |  \Large Optimal K: some titles
-author: |
- | \large Mena Whalen 
-  \vspace{-1.1mm}
-   \normalsize Department of Mathematics and Statistics \vspace{-1mm}
-  | \normalsize Loyola University Chicago \vspace{-1mm}
-   | \normalsize Chicago, IL 60660 \vspace{-1mm}
-   | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
-
-   \large Gregory J. Matthews 
-  \vspace{-1.1mm}
-   \normalsize Department of Mathematics and Statistics \vspace{-1mm}
-   \normalsize Loyola University Chicago \vspace{-1mm}
-   \normalsize Chicago, IL 60660 \vspace{-1mm}
-   \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
-
+title: "test"
+author: 
+  - "Angelos Vasilopoulos"
+  - "Gregory J. Matthews"
 abstract: |

### 2023-05-06 manuscript/manuscript.Rmd
“Saturday”

---
title: "test"
author:
- Angelos Vasilopoulos
- Gregory J. Matthews
abstract: "| Abstract \\vspace{2mm}\n| *Keywords*: Cross validation  \n"
output:
  pdf_document:
    df_print: kable
    number_sections: yes
    keep_tex: yes
  word_document: default
fontsize: 12pt
link-citations: yes
linkcolor: cyan
urlcolor: cyan
bibliography: /Users/angelos/Documents/research/optimal_k/optimal_k_references.bib
header-includes: \usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
indent: TRUE
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

\newpage

# Introduction {#sec:intro}

$k$-fold cross-validation is a popular method of error estimation for model selection in computational research.

n observations are divided into $k$ groups. In a first iteration, $i = k - 1$ groups are used as a training set. The remaining group is used as a test set to estimate prediction error. This process is repeated $i$ times, with a different test set in each iteration. The average of $k$ prediction errors ($PE$)
$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}PE_i$$

is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population.

Despite the popularity of cross-validation, there is limited focus in the literature on the question of what fold number $k$ is appropriate for cross-validation with a given data set. One popular idea is that selection of $k$ comes with a bias-variance trade-off, specifically, that as $k$ increases, the bias of k-fold error estimation
$$Bias(\hat{\theta}) = E(\hat{\theta}) - \theta$$
and variance of $k$-fold error estimation
$$Var(\hat{\theta}) = E(\hat{\theta}^2) - E(\hat{\theta})^2$$
as in
$$Err(\hat{\theta}) = Bias^2(\hat{\theta}) + Var(\hat{\theta})$$
decrease and increase, respectively. This idea appears in early literature (@Efron1983; @Kohavi2001).

Subsequent, albeit limited, literature argues differently. In the case of leave-one-out cross-validation, i.e., with $k$ = 1, previous literature suggests that asymptotically both bias and variance of error estimation decrease as $k$ increases (@Burman1989) and that bias and variance of error estimation are uniformly low (@Breiman1992).

More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error is not known, e.g., as mean-square stability (@Kale2011), which measures performance estimate variance across folds, or as loss stability (@Kumar2013), the proportion of times that a model outperforms competing models in all folds. However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross validation (@Bengio2004).

In addition to theoretical analysis of variance reduction by cross validation, there are some simulation results showing this phenomenon (@Zhang2015). However, simulations currently in the literature provide limited insight into the dependence of optimal fold number $k$ on sample size $n$. Here we present a simulation of linear regression and least absolute shrinkage and regression operator (LASSO) regression to observe the relationship of cross-validation fold number $k$ and model prediction error estimation accuracy for various sample sizes $n$.

# Simulation

Consider a population of size $N = 500,000$ with five of 100 features $X \sim N(0, 1)$ a linear combination
$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5 + \epsilon$$
where $\beta_1...\beta_5 = 1$ and $\epsilon \sim N(0, 10)$. Then the true model of the population is
$$E(Y) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5$$
and the mean squared error ($MSE$) of the true model is
$$MSE[E(Y)] = \theta = \frac{\sum_{i=1}^{N}[Y - E(Y)]^2}{N}.$$

From the population we take a sample of size $n$. We estimate $Y$ as $\hat{Y}$ by regression on a subset of $X$ and estimate $MSE(\hat{Y})$ as 
$$\hat{\theta} = \frac{\sum_{i=1}^{N}(Y - \hat{Y})^2}{N}$$
by $k$-fold cross-validation. We repeat this procedure in 1000 simulations for the same subsets of $X$, including the five-variable subset of the true model, and values of $k$. We then calculate simulation-wise $MSE(\hat{\theta})$ for each value of $k$.

We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression. To avoid data leakage, we introduce an inner 5-fold cross-validation loop for the selection of regularization parameter $\lambda$ as in the minimization of
$$\sum_{i=1}^{n}(Y_i - \sum_{j}X_{ij}\beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.$$

We perform the same 1000 simulations with $n$ = {100, 500, 1000} and consider as the optimal value of $k$ the value $k_{optimal}$ resulting in the lowest $MSE(\hat{\theta})$ for the true model the greatest number of times.

This is because in practice the model with the lowest prediction error is more likely to be selected. We refer to this as lowest-error model selection. As our simulation results show, however, it is possible for $k$-fold cross validation to result in calculations of $MSE(\hat{\theta})$ that suggest that a competing model has a lower prediction error than the true model. This false model may have the lowest prediction error in $k$-fold cross validation but its generalization error will be lower in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model, as $E(Y)$ is a linear combination of the true model only. Thus, it is preferable for the value of $k$ selected to result in the true model having the lowest estimated prediction error.

# Results

We find that for all $n$ as fold number $k$ increases, both bias and variance of error estimation decrease initially before leveling off (Figures 1 - 3).
```{r echo=FALSE}
source("~/Documents/research/optimal_k/results/optimal_k_plots.R")
figure1
figure2
figure3
```
We also find that for larger $n$, e.g., $n = 1000$, lowest-error model selection becomes less reliable as $k$ increases past a certain value, i.e., as $k$ increases past a certain value the true model has the lowest MSE less frequently. With increasing $n$ we also observe variability in the improvement of lowest-error model selection reliability (Figure 4; Table 1).
```{r echo=FALSE}
figure4
```
```{r}
kable_styling(kable(dffortable), latex_options = "hold_position")
```

# Conclusion

Early literature suggests that increasing cross-validation fold number is related to decreasing bias and increasing variance of error estimation. However, more recent work suggests that this is not the case. Instead, increasing $k$ results in bias and variance reduction. This phenomenon is observable in our simulation results, although it is also clear from the results that the reductions are not limitless, meaning that there is a point of diminishing returns in increasing $k$ and computational power allocation, as error estimation does not improve after a certain value of $k$.

Our results also indicate a relationship between optimal fold number $k$ and sample size $n$, although more data is necessary to elucidate a trend. This is interesting as a future direction, as modeling the relationship between $n$ and $k_{optimal}$ would have practical utility, potentially improving the selection of $k$ from a largely arbitrary decision between 5 and 10. Future research may also focus on error estimation of other models, including models capturing non-linear relationships or involving tuning of multiple hyper-parameters, e.g., random forest or gradient boosting.

# Acknowledgements {.unnumbered}

Dr. Gregory Matthews's Fall 2022 Predictive Analytics class.

# Supplementary Material {.unnumbered}

All code for reproducing the analyses in this paper is publicly available at <https://github.com/gjm112/optimal_k>.

# References

### 2023-05-06 manuscript/manuscript.tex
“Saturday”

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={test},
  pdfauthor={Angelos Vasilopoulos; Gregory J. Matthews},
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

\title{test}
\author{Angelos Vasilopoulos \and Gregory J. Matthews}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Abstract \vspace{2mm}\\
\emph{Keywords}: Cross validation
\end{abstract}

\newpage

\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

\(k\)-fold cross-validation is a popular method of error estimation for
model selection in computational research.

n observations are divided into \(k\) groups. In a first iteration,
\(i = k - 1\) groups are used as a training set. The remaining group is
used as a test set to estimate prediction error. This process is
repeated \(i\) times, with a different test set in each iteration. The
average of \(k\) prediction errors (\(PE\))
\[\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}PE_i\]

is meant to estimate the true model error \(\theta\), i.e., the error of
the model tested on the population.

Despite the popularity of cross-validation, there is limited focus in
the literature on the question of what fold number \(k\) is appropriate
for cross-validation with a given data set. One popular idea is that
selection of \(k\) comes with a bias-variance trade-off, specifically,
that as \(k\) increases, the bias of k-fold error estimation
\[Bias(\hat{\theta}) = E(\hat{\theta}) - \theta\] and variance of
\(k\)-fold error estimation
\[Var(\hat{\theta}) = E(\hat{\theta}^2) - E(\hat{\theta})^2\] as in
\[Err(\hat{\theta}) = Bias^2(\hat{\theta}) + Var(\hat{\theta})\]
decrease and increase, respectively. This idea appears in early
literature (Efron (\protect\hyperlink{ref-Efron1983}{1983}); Kohavi
(\protect\hyperlink{ref-Kohavi2001}{2001})).

Subsequent, albeit limited, literature argues differently. In the case
of leave-one-out cross-validation, i.e., with \(k\) = 1, previous
literature suggests that asymptotically both bias and variance of error
estimation decrease as \(k\) increases (Burman
(\protect\hyperlink{ref-Burman1989}{1989})) and that bias and variance
of error estimation are uniformly low (Breiman and Spector
(\protect\hyperlink{ref-Breiman1992}{1992})).

More recently have been proposed ways to quantify the variance reduction
achieved by cross-validation when the true prediction error is not
known, e.g., as mean-square stability (Kale, Kumar, and Vassilvitskii
(\protect\hyperlink{ref-Kale2011}{2011})), which measures performance
estimate variance across folds, or as loss stability (Kumar et al.
(\protect\hyperlink{ref-Kumar2013}{2013})), the proportion of times that
a model outperforms competing models in all folds. However, it has also
been demonstrated that, due to overlap between training and test sets in
cross-validation, there is no universal (i.e., valid under all
distributions) unbiased estimator of the variance of \(k\)-fold cross
validation (Bengio and Grandvalet
(\protect\hyperlink{ref-Bengio2004}{2004})).

In addition to theoretical analysis of variance reduction by cross
validation, there are some simulation results showing this phenomenon
(Zhang and Yang (\protect\hyperlink{ref-Zhang2015}{2015})). However,
simulations currently in the literature provide limited insight into the
dependence of optimal fold number \(k\) on sample size \(n\). Here we
present a simulation of linear regression and least absolute shrinkage
and regression operator (LASSO) regression to observe the relationship
of cross-validation fold number \(k\) and model prediction error
estimation accuracy for various sample sizes \(n\).

\hypertarget{simulation}{%
\section{Simulation}\label{simulation}}

Consider a population of size \(N = 500,000\) with five of 100 features
\(X \sim N(0, 1)\) a linear combination
\[Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5 + \epsilon\]
where \(\beta_1...\beta_5 = 1\) and \(\epsilon \sim N(0, 10)\). Then the
true model of the population is
\[E(Y) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5\]
and the mean squared error (\(MSE\)) of the true model is
\[MSE[E(Y)] = \theta = \frac{\sum_{i=1}^{N}[Y - E(Y)]^2}{N}.\]

From the population we take a sample of size \(n\). We estimate \(Y\) as
\(\hat{Y}\) by regression on a subset of \(X\) and estimate
\(MSE(\hat{Y})\) as
\[\hat{\theta} = \frac{\sum_{i=1}^{N}(Y - \hat{Y})^2}{N}\] by \(k\)-fold
cross-validation. We repeat this procedure in 1000 simulations for the
same subsets of \(X\), including the five-variable subset of the true
model, and values of \(k\). We then calculate simulation-wise
\(MSE(\hat{\theta})\) for each value of \(k\).

We perform an additional 1000 simulations predicting \(Y\) as
\(\hat{Y}\) by LASSO regression instead of linear regression. To avoid
data leakage, we introduce an inner 5-fold cross-validation loop for the
selection of regularization parameter \(\lambda\) as in the minimization
of
\[\sum_{i=1}^{n}(Y_i - \sum_{j}X_{ij}\beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.\]

We perform the same 1000 simulations with \(n\) = \{100, 500, 1000\} and
consider as the optimal value of \(k\) the value \(k_{optimal}\)
resulting in the lowest \(MSE(\hat{\theta})\) for the true model the
greatest number of times.

This is because in practice the model with the lowest prediction error
is more likely to be selected. We refer to this as lowest-error model
selection. As our simulation results show, however, it is possible for
\(k\)-fold cross validation to result in calculations of
\(MSE(\hat{\theta})\) that suggest that a competing model has a lower
prediction error than the true model. This false model may have the
lowest prediction error in \(k\)-fold cross validation but its
generalization error will be lower in the long-run (i.e., when tested on
a large part of the population) than the generalization error of the
true model, as \(E(Y)\) is a linear combination of the true model only.
Thus, it is preferable for the value of \(k\) selected to result in the
true model having the lowest estimated prediction error.

\hypertarget{results}{%
\section{Results}\label{results}}

We find that for all \(n\) as fold number \(k\) increases, both bias and
variance of error estimation decrease initially before leveling off
(Figures 1 - 3).

\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1} \end{center}

\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-2} \end{center}

\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-3} \end{center}

We also find that for larger \(n\), e.g., \(n = 1000\), lowest-error
model selection becomes less reliable as \(k\) increases past a certain
value, i.e., as \(k\) increases past a certain value the true model has
the lowest MSE less frequently. With increasing \(n\) we also observe
variability in the improvement of lowest-error model selection
reliability (Figure 4; Table 1).

\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{center}
\begin{table}[!h]
\centering
\begin{tabular}{r|r|r|r|r}
\hline
n & Minimum & Maximum & Range & Variance\\
\hline
100 & 240 & 306 & 66 & 140.0310\\
\hline
250 & 386 & 500 & 114 & 427.1585\\
\hline
500 & 619 & 813 & 194 & 676.5341\\
\hline
1000 & 861 & 931 & 70 & 711.2000\\
\hline
\end{tabular}
\end{table}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Early literature suggests that increasing cross-validation fold number
is related to decreasing bias and increasing variance of error
estimation. However, more recent work suggests that this is not the
case. Instead, increasing \(k\) results in bias and variance reduction.
This phenomenon is observable in our simulation results, although it is
also clear from the results that the reductions are not limitless,
meaning that there is a point of diminishing returns in increasing \(k\)
and computational power allocation, as error estimation does not improve
after a certain value of \(k\).

Our results also indicate a relationship between optimal fold number
\(k\) and sample size \(n\), although more data is necessary to
elucidate a trend. This is interesting as a future direction, as
modeling the relationship between \(n\) and \(k_{optimal}\) would have
practical utility, potentially improving the selection of \(k\) from a
largely arbitrary decision between 5 and 10. Future research may also
focus on error estimation of other models, including models capturing
non-linear relationships or involving tuning of multiple
hyper-parameters, e.g., random forest or gradient boosting.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

Dr.~Gregory Matthews's Fall 2022 Predictive Analytics class.

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

All code for reproducing the analyses in this paper is publicly
available at \url{https://github.com/gjm112/optimal_k}.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bengio2004}{}}%
Bengio, Yoshua, and Yves Grandvalet. 2004. {``No Unbiased Estimator of
the Variance of k-Fold Cross-Validation,''} June.

\leavevmode\vadjust pre{\hypertarget{ref-Breiman1992}{}}%
Breiman, Leo, and Philip Spector. 1992. {``Submodel Selection and
Evaluation in Regression. The x-Random Case.''} \emph{International
Statistical Review / Revue Internationale de Statistique} 60 (3): 291.
\url{https://doi.org/10.2307/1403680}.

\leavevmode\vadjust pre{\hypertarget{ref-Burman1989}{}}%
Burman, Prabir. 1989. {``A Comparative Study of Ordinary
Cross-Validation, v-Fold Cross-Validation and the Repeated
Learning-Testing Methods.''} \emph{Biometrika} 76 (September): 503--14.
\url{https://doi.org/10.1093/biomet/76.3.503}.

\leavevmode\vadjust pre{\hypertarget{ref-Efron1983}{}}%
Efron, Bradley. 1983. {``Estimating the Error Rate of a Prediction Rule:
Improvement on Cross-Validation.''} \emph{Journal of the American
Statistical Association} 78 (382): 316--31.
\url{https://doi.org/10.1080/01621459.1983.10477973}.

\leavevmode\vadjust pre{\hypertarget{ref-Kale2011}{}}%
Kale, Satyen, Ravi Kumar, and Sergei Vassilvitskii. 2011.
{``Cross-Validation and Mean-Square Stability.''} In, 487--95.

\leavevmode\vadjust pre{\hypertarget{ref-Kohavi2001}{}}%
Kohavi, Ron. 2001. {``A Study of Cross-Validation and Bootstrap for
Accuracy Estimation and Model Selection''} 14 (March).

\leavevmode\vadjust pre{\hypertarget{ref-Kumar2013}{}}%
Kumar, Ravi, Daniel Lokshtanov, Sergei Vassilvitskii, and Andrea
Vattani. 2013. {``Near-Optimal Bounds for Cross-Validation via Loss
Stability.''} In \emph{Proceedings of the 30th International Conference
on Machine Learning}, edited by Sanjoy Dasgupta and David McAllester,
28:27--35. Proceedings of Machine Learning Research 1. Atlanta, Georgia,
USA: PMLR. \url{https://proceedings.mlr.press/v28/kumar13a.html}.

\leavevmode\vadjust pre{\hypertarget{ref-Zhang2015}{}}%
Zhang, Yongli, and Yuhong Yang. 2015. {``Cross-Validation for Selecting
a Model Selection Procedure.''} \emph{Journal of Econometrics} 187 (1):
95--112. \url{https://doi.org/10.1016/j.jeconom.2015.02.006}.

\end{CSLReferences}

\end{document}

### 2023-05-06 plots/optimal_k_plots.R

@@ -1,0 +1,158 @@
+
+# load libraries
+library(ggplot2)
+library(ggpubr)
+library(kableExtra)
+
+# create population
+N <- 500000
+x<-data.frame(matrix(nrow=500000,ncol=100))
+set.seed(2022)
+n_var <- 1:100
+for(i in n_var){
+  x[,i]<-rnorm(n=N,mean=0,sd=1)
+}
+
+# define parameters
+beta0 <- 0
+beta1 <- 1
+beta2 <- 1
+beta3 <- 1
+beta4 <- 1
+beta5 <- 1
+sigma <- 10
+
+# define theta
+Ey <- beta0 + beta1*x[,1] + beta2*x[,2] + beta3*x[,3] + beta4*x[,4] + beta5*x[,5]
+y <- Ey + rnorm(nrow(x),0,sigma)
+theta <- mean((y - Ey)^2)
+
+# choose models
+variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
+
+# loop settings
+n <- c(100, 250, 500, 1000)
+nsim <- 1000
+k <- list(n100 = seq(2, 100, 2),
+          n250 = c(2, seq(10, 250, 10)),
+          n500 = c(2, seq(10, 500, 10)),
+          n1000 = c(2, 10, 100, 500, 1000)) # for n = 1000
+n_mod <- length(variables)
+out <- list(n100 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
+            n250 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
+            n500 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
+            n1000 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output.rds"))
+titles <- list("n = 100", "n = 250", "n = 500", "n = 1000")
+plotsout <- list(list(), list(), list())
+figures <- list()
+
+for(i in 1:length(out)){
+  # read results
+  theta.hat <- mse.hat <- array(unlist(out[[i]]), dim = c(length(k[[i]]), n_mod, nsim))
+  
+  # biassq + var = mse
+  biassq <- as.data.frame((apply(theta.hat, c(1, 2), mean) - theta)^2)
[…]

### 2023-05-19 manuscript/manuscript copy.Rmd
May nineteenth

---
title: "Simulations of fold number selection in cross-validation of linear and LASSO regression"
author:
- Angelos Vasilopoulos
- Gregory J. Matthews
abstract: "*k*-fold cross-validation is a popular method of error estimation for model selection in computational research. However, there is limited focus in the literature on the question of what fold number *k* is appropriate for cross-validation with a dataset of a given size. Here we review the literature on this topic and present a simulation of linear and least absolute shrinkage and selection operator (LASSO) regression prediction error estimation at various values of *k* and sample size *n*. In agreement with a growing body of literature, we find that contrary to a persisting understanding, there is no bias-variance trade-off in selection of *k*. | *Keywords*: cross-validation, optimization, fold number"
output:
  pdf_document:
    df_print: kable
    number_sections: yes
    keep_tex: yes
  word_document: default
fontsize: 12pt
link-citations: yes
linkcolor: cyan
urlcolor: cyan
bibliography: /Users/angelos/Documents/research/optimal_k/manuscript/optimal_k_references.bib
header-includes: \usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
indent: TRUE
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	fig.pos = 'H'
)
gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

# Introduction {#sec:intro}

$k$-fold cross-validation is a popular method of error estimation for model selection in computational research. $n$ observations are divided into $k$ groups. In a first iteration, $i = k - 1$ groups are used as a training set. The remaining group is used as a test set to calculate estimated prediction error ($\mathrm{\widehat{PE}}$). This process is repeated $i$ times, with a different test set in each iteration. The average of $k$ $\mathrm{\widehat{PE}}$
$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i$$
is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population.

Despite the popularity of cross-validation, there is limited focus in the literature on the question of what fold number $k$ is appropriate for cross-validation with a dataset of a given size. One popular idea is that selection of $k$ comes with a bias-variance trade-off—specifically, that as $k$ increases, the bias $Bias(\hat{\theta}) = \mathrm{E}(\hat{\theta}) - \theta$ and variance $Var(\hat{\theta}) = \mathrm{E}(\hat{\theta}^2) - \mathrm{E}(\hat{\theta})^2$ of $k$-fold error estimation,
as in $Err(\hat{\theta}) = Bias^2(\hat{\theta}) + Var(\hat{\theta})$, decrease and increase, respectively. This idea appears in early literature (@Efron1983; @Kohavi2001).

Subsequent, albeit limited, literature argues differently. In the case of leave-one-out cross-validation (LOOCV), i.e., with $k = n$, some authors suggest that asymptotically both bias and variance of error estimation decrease as $k$ increases (@Burman1989) and that bias and variance of error estimation are uniformly low (@Breiman1992).

More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011), where
$$\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,$$
i.e., the expectation of the squared difference in loss $\ell$ of an algorithm $\mathcal A$ over a dataset $S$ and over a dataset $S'$ such that one element $z' \sim Z$ has been added to $S$ is less than $\beta$, which depends on the algorithm; or as loss stability (@Kumar2013), where
$$\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,$$
$\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)$, and $\gamma$ depends on the algorithm. However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross-validation (@Bengio2004).

In addition to theoretical analyses of variance reduction by cross-validation, there are some simulation results showing this phenomenon. However, simulations currently in the literature provide limited insight into the dependence of optimal fold number $k_{\mathrm{optimal}}$ on sample size $n$ (@Zhang2015) or involve biased variance calculations (@Marcot2021). Here we present a simulation of linear regression and least absolute shrinkage and selection operator (LASSO) regression to observe the relationship of cross-validation fold number $k$ to model selection accuracy for various samples of size $n$ of a known population.

# Simulation

Consider a population of size N = 500,000 with five of $P = 100$ features $X_1...X_{100} \sim N(0, 1)$ a linear combination of feature $Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon$
where $\beta_1 = \cdots = \beta_5 = 1$ and $\epsilon \sim N(0, 10)$. The true model of $Y$ is $\mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5$
and the mean squared error ($\mathrm{MSE}$) of the true model is
$$\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$

From the population we take a sample of size $n$. We estimate $Y$ as $\hat{Y}$ by regression on a subset of $X_1...X_{100}$, regression coefficients estimated by the least squares method, and compute $\mathrm{MSE}(\hat{Y})$ by $k$-fold cross-validation as
$$\hat{\theta} = \mathrm{MSE}(\hat{Y}) = \frac{1}{k}\sum_{j=1}^{k}\frac{\sum_{i=1}^{n/k}(Y_{j_i} - \hat{Y_{j_i}})^2}{n/k}$$
where $j_i$ is the index of the $i^{th}$ element of the $j^{th}$ fold.

In addition to regression with $X_1...X_5$, the true model, we under-fit with $X_1...X_3$, over-fit with $X_1...X_{20}$ and $X_1...X_{100}$, and do regression with noise $X_6...X_{100}$ only for each $k \in$ {2, 10, 20, 30, ..., $n \in$ {100, 200, 300, ..., 1000}}. We perform this simulation 1000 times and then calculate simulation-wise $\mathrm{MSE}(\hat{\theta})$ for each $k$, for each $n$.

We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression for each $k \in$ {2, 10, 20, 30, ..., $n \in$ {100, 250, 500, 750, 1000}}, due to computational limitations. To avoid data leakage, we introduce an inner 5-fold cross-validation loop for the selection of parameter $\lambda$ as in the minimization of
$$\sum_{i=j}^{p}(Y_j - \beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.$$

For each $k$, for each $n$, we count the number of times $C$ that the true model is selected and consider as optimal fold number for each $n$ the value $k_{\mathrm{optimal}}$, at which $C$ is the highest.

This is because in practice the model with the lowest $\mathrm{\widehat{PE}}$ is more likely to be selected. We refer to this as lowest-error model selection. As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model has a lower $\mathrm{\widehat{PE}}$ than the true model. This false model may have the lowest $\mathrm{\widehat{PE}}$ but its generalization error will be higher in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model. Thus, it is preferable for the value of $k$ selected to result in the true model having the lowest $\mathrm{\widehat{PE}}$.

Having considered the effect of $k$ on model selection accuracy for a given $n$, we may also consider the relationship of $k_{\mathrm{optimal}}$ and $n$. At this point it is important to take into account the possibility of diminishing returns after a certain value of $k$: that $C$ appears to increase until $k = n$ for certain $n$ does not mean that for those $n$ LOOCV results in significantly higher $C$. In other words, after a certain point, improvements in $C$ become negligible and resource-expensive increases of $k$ become unnecessary. We refer to this "point of diminishing returns" as $k_{optimal}^*$.

To estimate $k_{optimal}^*$ for each $n$, we first fit to the dataset $D(k,C)$ for each sample slize $n$ a non-linear least squares model of the form
$$\hat{C}(k) = \frac{A}{k} + B$$
with constants $A$ and $B$ specific to each $n$. We then perform $n/10 - 1$ Mann-Kendall tests of trend significance on the values fitted at $k_i...n \in k$, $i = [1..(n/10 - 1)]$, as the Mann-Kendall test requires at least three data points. We apply a Bonferroni correction to account for the problem of multiple comparisons, with $\alpha = 0.05$ and a corrected cut-off of $p* = \alpha/(n/10 - 1)$. Then 
$$k_{optimal}^* = \max_{p < p^*}k$$
since $\hat{C}(k)$ increases monotonically.

# Results and discussion

Interpretations of early literature have resulted in lasting misconceptions about the use of cross-validation. Such misconceptions include the idea that there is a bias-variance trade-off
$$Bias^2(\hat{\theta}) \propto \frac{1}{Var(\hat{\theta})}$$
associated with selection of $k$ and that $k = 10$ is the best value to use in $k$-fold cross-validation (@Kohavi2001).

In agreement with a small but growing body of literature, our simulation results suggest that neither of these ideas are necessarily correct. Instead, in the context of linear and LASSO regression with normally distributed data, we find that for various $n$ both bias and variance decrease as $k$ increases (Figures 1 - 3), i.e., $Bias^2(\hat{\theta}) \propto Var(\hat{\theta})$, and although 10-fold cross-validation seems to be superior for large $n$, for smaller samples other values of $k$ are better for model selection (Figure 4).

As previous authors have noted, it is important to distinguish the possible goals of cross-validation: estimation of $\mathrm{PE}$ and model selection (@Zhang2015), which previously have been conflated (@Kohavi2001). In this study, as described in the previous section, we define $k_{\mathrm{optimal}}$ in terms of its utility in the latter. For LASSO regression, we  find that for larger $n$, e.g., $n = 1000$, as $k_{\mathrm{optimal}}$ increases past a certain value lowest-error model selection becomes less reliable, i.e., $C$ decreases, i.e., the true model has the lowest $\mathrm{MSE}$ less frequently (Figure 4).

The reason for this is related to what is known as the cross-validation paradox (@Yang2006): greater quantity of data results in more accurate estimation of $\mathrm{PE}$; occasionally, this makes model-wise differences in $\mathrm{\widehat{PE}} = CV_{(k)}$ less exaggerated, making it more difficult to distinguish between models and resulting in less accurate model selection.

However, we do not observe the same for linear regression in our simulations. Although cross-validation results in similar bias-variance reduction in both cases, in the case of linear regression $C$ increases with increasing $k$, so that LOOCV appears to result in the most accurate model selection. However, this discrepancy is not surprising, as our linear regression simulation does not involve variable selection, so that the differences in the models are more pronounced. To observe the cross-validation paradox in linear regression would require comparison of more similar linear models predisposed to similarity in $\widehat{\mathrm{PE}}$, as in the simulation of LASSO regression.

The results dispalyed in Figure 5 suggest an increase in $k_{\mathrm{optimal}}/n$ with increasing $n$ (Figure 5). In other words, as $n$ increases, the training set size required for optimal model selection accuracy increases.

# Conclusion

Early literature suggests that increasing cross-validation fold number is related to decreasing bias and increasing variance of error estimation. However, more recent work suggests that this is not the case. Instead, increasing $k$ results in bias and variance reduction. This phenomenon is observable in our simulation results, although it is also clear from the results that the reductions are not limitless, meaning that there is a point of diminishing returns in increasing $k$ and computational power allocation, as error estimation does not improve, at least not substantially, after a certain value of $k$.

Our results also indicate a relationship between $k_{\mathrm{optimal}}$ and sample size $n$, although more data is necessary to elucidate a trend. This is interesting as a future direction, as modeling the relationship between $n$ and $k_{\mathrm{optimal}}$ would have practical utility, potentially improving the selection of $k$ from a largely arbitrary decision between 5 and 10.

Future research may also focus on error estimation of other models, including models capturing non-linear relationships or involving tuning of multiple hyper-parameters, e.g., random forest or gradient boosting. It may also be interesting to study the value of $k_{\mathrm{optimal}}$ in repeated cross-validation or nested cross-validation, with the value of $k$ variable in the inner loop, outer loop, or both.

# Appendix

```{r echo=FALSE, fig.cap = "Squared bias vs. fold size k for sample sizes n. As k increases, bias decreases initially before leveling off."}
source("~/optimal_k_git/plots/optimal_k_plots.R")
figure1
```

```{r echo=FALSE, fig.cap = "Variance vs. fold size k for sample sizes n. As k increases, variance decreases initially before leveling off."}
figure2
```

```{r echo=FALSE, fig.cap = "Mean squared error (MSE) vs. fold size k for sample sizes n. As k increases, MSE decreases initially before leveling off."}
figure3
```

```{r echo=FALSE, fig.cap = "Count of true model selection by the lowest-error criterion vs. fold size k for sample sizes n."}
figure4
```

```{r eval=FALSE, include=FALSE}
kable_styling(kable(dffortable, booktabs = TRUE, caption = "Table 1. Minimum, maximum, range, and variance of true model selection counts per fold number k for various sample sizes n."), latex_options = "hold_position")
```

# Acknowledgements {.unnumbered}

Dr. Gregory Matthews's Fall 2022 Predictive Analytics class.

# Supplementary Material {.unnumbered}

All code for reproducing the analyses in this paper is publicly available at <https://github.com/gjm112/optimal_k>.

# References

### 2023-05-19 plots/optimal_k_plots.R

@@ -38,10 +38,10 @@ k <- list(n100 = seq(2, 100, 2),
           n500 = c(2, seq(10, 500, 10)),
           n1000 = c(2, 10, 100, 500, 1000)) # for n = 1000
 n_mod <- length(variables)
-out <- list(n100 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
-            n250 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
-            n500 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
-            n1000 = readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output.rds"))
+out <- list(n100 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
+            n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
+            n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
+            n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
 titles <- list("n = 100", "n = 250", "n = 500", "n = 1000")
 plotsout <- list(list(), list(), list())
 figures <- list()

### 2023-05-19 simulations/simulation_lasso_parallel.R

@@ -27,6 +27,7 @@ variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 # loop settings
 n <- 750 # 250 # 500 # 100 # 1000
 nsim <- 1000
+# k <- c(seq(2, 100, 2), 500, 1000)
 # k <- seq(2, 100, 2) # for n = 100 and n = 1000
 # k <- c(2, seq(10, 250, 10)) # for n = 250
 k <- c(2, seq(10, 500, 10)) # for n = 500
@@ -70,7 +71,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("~/Documents/research/optimal_k/results/parallel_simulation_lasso_output_k_2_500_n_500.rds")
+out <- readRDS("~/optimal_k_git/results/parallel_simulation_lasso_output_k_2_500_n_500 copy.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -115,10 +116,11 @@ for(sim in 1:nsim){
 }
 corrects <- colSums(results == 2)
 plot(k, corrects, type = "b", ylab = "How many times correct model had lowest MSE (n = 250)")
+df <- data.frame(k, corrects)
 
+# plot
 samplesize <- c(100, 250, 500, 1000)
-optimalk <- c(24, 80, 30, 4)
-
+optimalk <- c(24, 80, 70, 4)
 abline(plot(samplesize, optimalk))
 
 

### 2023-05-19 simulations/simulation_linear_non-parallel.R

@@ -1,0 +1,85 @@
+# create population
+N <- 500000
+x<-data.frame(matrix(nrow=500000,ncol=100))
+set.seed(2022)
+n_var <- 1:100
+for(i in n_var){
+  x[,i]<-rnorm(n=N,mean=0,sd=1)
+}
+
+# define parameters
+beta0 <- 0
+beta1 <- 1
+beta2 <- 1
+beta3 <- 1
+beta4 <- 1
+beta5 <- 1
+sigma <- 10
+
+# define theta
+Ey <- beta0 + beta1*x[,1] + beta2*x[,2] + beta3*x[,3] + beta4*x[,4] + beta5*x[,5]
+y <- Ey + rnorm(nrow(x),0,sigma)
+theta <- mean((y - Ey)^2)
+
+# choose models
+variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
+
+# c(1,2),c(1:3),c(1:4),c(1:5),c(1:10),
+# c(1:20),c(1:35),c(1:55),c(1:100),c(1,6:15),
+# c(1,6:50),c(1,2,6:15),c(1,2,6:50),c(1:3,6:15),c(1:3,6:50),
+# c(1:4,6:15),c(1:4,6:50),c(6:20),c(6:30),c(6:45),
+# c(6:70),c(6:100))
+
+# loop settings
+nsim <- 1000
+n <- 100 # 1000
+k <- c(2, seq(10, 100, 10)) # c(2, 10, 100, 500, 1000)
+n_mod <- length(variables)
+
+# create objects
+results <- array(NA, dim = c(nsim, length(k), n_mod), dimnames = list(rep("sim", nsim), rep("k", length(k)), rep("mod", n_mod)))
+mod <- list()
+y_hat <- matrix(nrow = n, ncol = n_mod)
+
+# loop 1
+set.seed(2022)
+for(j in 1:nsim){
+  ind <- sample(1:N, n, replace = FALSE)
+  for(l in 1:length(k)){ # choose lth fold number to test
+    fold <- sample(1:k[l], n, replace = TRUE)
+    for(i in 1:k[l]){ # outer cv loop
+      for(m in 1:n_mod){print(c(j, l, i, m))
+        mod[[m]] <- lm(y ~ ., cbind(x[ind, variables[[m]]][fold != i,], y = y[ind][fold != i]))
+        y_hat[fold == i, m] <- predict(mod[[m]], x[ind, variables[[m]]][fold == i, ])
+      }
+    }
[…]

### 2023-05-19 simulations/simulation_linear_parallel.R

@@ -1,3 +1,4 @@
+
 # create population
 N <- 500000
 x<-data.frame(matrix(nrow=500000,ncol=100))
@@ -25,9 +26,9 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- 1000
+n <- 500 # 1000
 nsim <- 1000
-k <- c(2, 10, 100, 500, 1000)
+k <- c(2, seq(10, n, 10)) # c(2, 10, 100, 500, 1000)
 n_mod <- length(variables)
 
 # create objects
@@ -64,8 +65,8 @@ out <- foreach(j = 1:nsim, .combine = "list") %dopar% {
 }
 
 # read results
-out <- readRDS("~/Documents/research/optimal_k/results/parallel_simulation_linear_output.rds")
-theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
+out <- readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_500_n_500.rds")
+theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim), dimnames = list(rep("k", length(k)), rep("mod", n_mod), rep("sim", nsim)))
 
 # biassq + var = mse
 biassq <- (apply(theta.hat, c(1, 2), mean) - theta)^2
@@ -75,8 +76,8 @@ mse <- apply((theta.hat - theta)^2, c(1, 2), mean)
 # plots
 library(plotrix)
 
-# bias^2 plot
-plot(k, biassq[, 1], type = "b", ylim = c(0, 400), ylab = "bias^2")
+# bias^2 plots
+plot(k, biassq[, 1], type = "b", ylim = c(min(biassq), max(biassq)), ylab = "bias^2")
 points(k, biassq[, 2], type = "b", col = "blue")
 points(k, biassq[, 3], type = "b", col = "red")
 points(k, biassq[, 4], type = "b", col = "purple")
@@ -85,7 +86,7 @@ legend(600, 250, legend=c("1:3", "1:5", "1:20", "1:100", "6:100"),
        col=c("black", "blue", "red", "purple", "green"), lty = 1, cex = 0.8)
 
 # variance plot
-plot(k, var[, 1], type = "b", ylim = c(10, 70), ylab = "var")
[…]

### 2023-05-31 manuscript/manuscript.Rmd
“May31”

@@
 ---
-title: "test"
+title: "Simulations of fold number selection in cross-validation of linear and LASSO regression"
 author:
@@
 - Gregory J. Matthews
-abstract: "| Abstract \\vspace{2mm}\n| *Keywords*: Cross validation  \n"
+abstract: "*k*-fold cross-validation is a popular method of error estimation for model selection in computational research.
+However, there is limited focus in the literature on the question of what fold number *k* is appropriate for various dataset dimensions.
+Here we review relevant literature and present a simulation of linear and least absolute shrinkage and selection operator (LASSO) regression prediction error estimation at various values of *k* and sample size *n*.
+In agreement with a growing body of literature, we find that contrary to a persisting understanding, there is no bias-variance trade-off in selection of *k*.
+Instead, with increasing *k* both bias and variance decrease, perhaps asymptotically.
+Our results also suggest a predictable relationship between optimal values of *k* and *n*.
+| *Keywords*: cross-validation, optimization, fold number"
 output:
@@
 urlcolor: cyan
-bibliography: /Users/angelos/Documents/research/optimal_k/optimal_k_references.bib
-header-includes: \usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
+bibliography: /Users/angelos/Documents/research/optimal_k/manuscript/optimal_k_references.bib
+header-includes: \usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t} \usepackage[labelsep=period]{caption}
 indent: TRUE
@@
 	message = FALSE,
-	warning = FALSE
+	warning = FALSE,
+	fig.pos = 'H'
 )
@@
 
-\newpage
-
 # Introduction {#sec:intro}
@@
 $k$-fold cross-validation is a popular method of error estimation for model selection in computational research.
-
-n observations are divided into $k$ groups.
+$n$ observations are divided into $k$ groups.
 In a first iteration, $i = k - 1$ groups are used as a training set.
-The remaining group is used as a test set to estimate prediction error.
+The remaining group is used as a test set to calculate estimated prediction error ($\mathrm{\widehat{PE}}$).
 This process is repeated $i$ times, with a different test set in each iteration.
-The average of $k$ prediction errors ($PE$)
-$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}PE_i$$
-
+The average of $k$ estimates of $\mathrm{PE}$
+$$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i$$
 is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population.
 
-Despite the popularity of cross-validation, there is limited focus in the literature on the question of what fold number $k$ is appropriate for cross-validation with a given data set.
-One popular idea is that selection of $k$ comes with a bias-variance trade-off, specifically, that as $k$ increases, the bias of k-fold error estimation
-$$Bias(\hat{\theta}) = E(\hat{\theta}) - \theta$$
-and variance of $k$-fold error estimation
-$$Var(\hat{\theta}) = E(\hat{\theta}^2) - E(\hat{\theta})^2$$
-as in
-$$Err(\hat{\theta}) = Bias^2(\hat{\theta}) + Var(\hat{\theta})$$
-decrease and increase, respectively. This idea appears in early literature (@Efron1983; @Kohavi2001).
+Despite the popularity of cross-validation, there is limited focus in the literature on the question of what fold number $k$ is appropriate for cross-validation with a dataset of a given size.
+One popular idea is that selection of $k$ comes with a bias-variance trade-off—specifically, that as $k$ increases, the bias $Bias(\hat{\theta}) = \mathrm{E}(\hat{\theta}) - \theta$ and variance $Var(\hat{\theta}) = \mathrm{E}(\hat{\theta}^2) - \mathrm{E}(\hat{\theta})^2$ of $k$-fold error estimation decrease and increase, respectively.
+This idea appears in early literature (@Efron1983; @Kohavi2001) but also in modern, widely used textbooks (@James2021).
 
 Subsequent, albeit limited, literature argues differently.
-In the case of leave-one-out cross-validation, i.e., with $k$ = 1, previous literature suggests that asymptotically both bias and variance of error estimation decrease as $k$ increases (@Burman1989) and that bias and variance of error estimation are uniformly low (@Breiman1992).
+In the case of leave-one-out cross-validation (LOOCV), i.e., with $k = n$, some authors suggest that asymptotically both bias and variance of error estimation decrease as $k$ increases (@Burman1989) and that bias and variance of error estimation are uniformly low (@Breiman1992).
 
-More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error is not known, e.g., as mean-square stability (@Kale2011), which measures performance estimate variance across folds, or as loss stability (@Kumar2013), the proportion of times that a model outperforms competing models in all folds.
-However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross validation (@Bengio2004).
+More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011), where
+$$\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,$$
+i.e., the expectation of the squared difference in loss $\ell$ of an algorithm $\mathcal A$ over a dataset $S$ and over a dataset $S'$ such that one element $z' \sim Z$ has been added to $S$ is less than $\beta$, which depends on the algorithm; or as loss stability (@Kumar2013), where
+$$\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,$$
+$\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)$, and $\gamma$ depends on the algorithm.
+However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross-validation (@Bengio2004).
 
-In addition to theoretical analysis of variance reduction by cross validation, there are some simulation results showing this phenomenon (@Zhang2015).
-However, simulations currently in the literature provide limited insight into the dependence of optimal fold number $k$ on sample size $n$.
-Here we present a simulation of linear regression and least absolute shrinkage and regression operator (LASSO) regression to observe the relationship of cross-validation fold number $k$ and model prediction error estimation accuracy for various sample sizes $n$.
+In addition to theoretical analyses of variance reduction by cross-validation, there are some simulation results showing this phenomenon.
+However, simulations currently in the literature provide limited insight into the dependence of optimal fold number $k_\mathrm{optimal}$ on sample size $n$ (@Zhang2015) or involve biased variance calculations (@Marcot2021).
+Here we present a simulation of linear regression and least absolute shrinkage and selection operator (LASSO) regression to observe the relationship of cross-validation fold number $k$ to model selection accuracy for various samples of size $n$ of a known population.
 
@@
 
-Consider a population of size $N = 500,000$ with five of 100 features $X \sim N(0, 1)$ a linear combination
-$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5 + \epsilon$$
-where $\beta_1...\beta_5 = 1$ and $\epsilon \sim N(0, 10)$. Then the true model of the population is
-$$E(Y) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5$$
-and the mean squared error ($MSE$) of the true model is
-$$MSE[E(Y)] = \theta = \frac{\sum_{i=1}^{N}[Y - E(Y)]^2}{N}.$$
+Consider a population of size N = 500,000 with five of $P = 100$ features $X_1...X_{100} \sim N(0, 1)$ a linear combination of feature $Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon$
+where $\beta_1 = \cdots = \beta_5 = 1$ and $\epsilon \sim N(0, 10)$. The true model of $Y$ is $f \in F$, a set of competing models. $f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5$
+and its mean squared error ($\mathrm{MSE}$) is
+$$\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$
 
-From the population we take a sample of size $n$. We estimate $Y$ as $\hat{Y}$ by regression on a subset of $X$ and estimate $MSE(\hat{Y})$ as 
-$$\hat{\theta} = \frac{\sum_{i=1}^{N}(Y - \hat{Y})^2}{N}$$
-by $k$-fold cross-validation.
-We repeat this procedure in 1000 simulations for the same subsets of $X$, including the five-variable subset of the true model, and values of $k$.
-We then calculate simulation-wise $MSE(\hat{\theta})$ for each value of $k$.
+From the population we take a sample of size $n$.
+We estimate $Y$ as $\hat{Y}$ by regression on a subset of $X_1...X_{100}$, regression coefficients estimated by the least squares method, and compute $\mathrm{MSE}(\hat{Y})$ by $k$-fold cross-validation as
+$$\hat{\theta} = \mathrm{MSE}(\hat{Y}) = \frac{1}{k}\sum_{j=1}^{k}\frac{\sum_{i=1}^{n/k}(Y_{j_i} - \hat{Y_{j_i}})^2}{n/k}$$
+where $j_i$ is the index of the $i^{th}$ element of the $j^{th}$ fold.
 
-We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression.
-To avoid data leakage, we introduce an inner 5-fold cross-validation loop for the selection of regularization parameter $\lambda$ as in the minimization of
-$$\sum_{i=1}^{n}(Y_i - \sum_{j}X_{ij}\beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.$$
+In addition to regression with $X_1...X_5$, $f$, we under-fit with $X_1...X_3$, over-fit with $X_1...X_{20}$ and $X_1...X_{100}$, and do regression with noise $X_6...X_{100}$ only, for each $k \in$ {2, 10, 20, 30, ..., $n$} for each  $n \in$ {100, 200, 300, ..., 1000}.
+We perform this simulation 1000 times and then calculate simulation-wise $\mathrm{MSE}(\hat{\theta})$ for each $k$, for each $n$.
 
-We perform the same 1000 simulations with $n$ = {100, 500, 1000} and consider as the optimal value of $k$ the value $k_{optimal}$ resulting in the lowest $MSE(\hat{\theta})$ for the true model the greatest number of times.
+We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression for each $k \in$ {2, 10, $n$/10, 2$n$/10, 3$n$/10, ..., $n$} for each $n \in$ {250, 500, 750, 1000}.
+To avoid data leakage, we introduce an inner 5-fold cross-validation loop for the selection of parameter $\lambda$ as in the minimization of
+$$\sum_{i=j}^{p}(Y_j - \beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.$$
 
-This is because in practice the model with the lowest prediction error is more likely to be selected.
+For each $k$, for each $n$, we count the number of times $A$ that the true model is selected and consider as optimal fold number for each $n$ the value $k_\mathrm{optimal}$, at which $A$ is the highest.
+
+This is because in practice the model with the lowest $\mathrm{\widehat{PE}}$ is more likely to be selected.
 We refer to this as lowest-error model selection.
-As our simulation results show, however, it is possible for $k$-fold cross validation to result in calculations of $MSE(\hat{\theta})$ that suggest that a competing model has a lower prediction error than the true model.
-This false model may have the lowest prediction error in $k$-fold cross validation but its generalization error will be lower in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model, as $E(Y)$ is a linear combination of the true model only.
-Thus, it is preferable for the value of $k$ selected to result in the true model having the lowest estimated prediction error.
+As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model $\hat{f}$ has a lower $\mathrm{\widehat{PE}}$ than the true model $f$.
+This false model may have the lowest $\mathrm{\widehat{PE}}$ but its generalization error will be higher in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model.
+Thus, it is preferable for the value of $k$ selected to result in $f$ having the lowest $\mathrm{\widehat{PE}}$.
 
-# Results
+Having considered the effect of $k$ on model selection accuracy for a given $n$, we may also consider the relationship of $k_\mathrm{optimal}$ and $n$.
+Our results suggest that after a certain value of $k$, changes in $A$ become negligible, in the case of linear regression, or negative, in the case of LASSO regression, and resource-expensive increases of fold number become undesirable.
+We refer to this "point of diminishing returns" as $k_\mathrm{optimal}^*$.
 
-We find that for all $n$ as fold number $k$ increases, both bias and variance of error estimation decrease initially before leveling off (Figures 1 - 3).
-```{r echo=FALSE}
-source("~/Documents/research/optimal_k/results/optimal_k_plots.R")
-figure1
-figure2
-figure3
-```
-We also find that for larger $n$, e.g., $n = 1000$, lowest-error model selection becomes less reliable as $k$ increases past a certain value, i.e., as $k$ increases past a certain value the true model has the lowest MSE less frequently.
-With increasing $n$ we also observe variability in the improvement of lowest-error model selection reliability (Figure 4; Table 1).
-```{r echo=FALSE}
-figure4
-```
-```{r}
-kable_styling(kable(dffortable), latex_options = "hold_position")
-```
+To estimate $k_\mathrm{optimal}*$ for each $n$, we first fit to the dataset $(k,A)$ for each sample size $n$ a non-linear least squares model of the form
+$$\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}$$
+with constants $B$ to $E$ specific to each $n$ and constants $B$ and $E$ set to zero for linear regression.
+We then draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that $k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}$.
+
+# Results and discussion
+
+Interpretations of early literature have resulted in lasting misconceptions about the use of cross-validation.
+Such misconceptions include the idea that there is a bias-variance trade-off $Bias^2(\hat{\theta}) \propto 1/Var(\hat{\theta})$ associated with selection of $k$ and that $k = 10$ is the best value to use in $k$-fold cross-validation (@Kohavi2001).
+
+In agreement with a small but growing body of literature, our simulation results suggest that neither of these ideas are necessarily correct.
+Instead, in the context of linear and LASSO regression with standard normal data and certain random error, we find that for various $n$ both bias and variance decrease as $k$ increases (Figures 1 - 6), i.e., $Bias^2(\hat{\theta}) \propto Var(\hat{\theta})$, and although in the case of LASSO, 10-fold cross-validation seems to be a near-optimal choice for large $n$, for smaller samples and linear regression other values of $k$ appear to be optimal for model selection (Figures 7 and 8).
+
+As previous authors have noted (@Zhang2015), it is important to distinguish between possible goals of cross-validation that previously have been conflated (@Kohavi2001): estimation of $\mathrm{PE}$, in which case $k_\mathrm{optimal} = {\arg\min}_{k}(\mathrm{PE} - \mathrm{\widehat{PE}})$, and model selection, in which case $$k_\mathrm{optimal} = {\arg\max}_k(A)$$, which is related to the definition of $k_\mathrm{optimal}^*$ used in this study.
+
+We find that in the cases of both linear and LASSO regression, as $n$ increases, $k_\mathrm{optimal}^*$ increases but at a lower rate than $n$, such that $k_\mathrm{optimal}^*/n$ decreases, perhaps asymptotically.
+For LASSO regression, we also find that as $k$ increases past a certain value, lowest-error model selection becomes less reliable, i.e., $\hat{A}$ decreases, i.e., the true model has the lowest $\mathrm{MSE}$ less frequently (Figure 8).
+The rate of this reduction increases with $n$.
+
+The reason for this is related to what is known as the cross-validation paradox (@Yang2006): greater quantity of data results in more accurate estimation of $\mathrm{PE}$; occasionally, this makes model-wise differences in $\mathrm{\widehat{PE}} = CV_{(k)}$ less exaggerated, making it more difficult to distinguish between models and resulting in less accurate model selection.
+
+However, we do not observe the cross-validation paradox in our linear regression simulations.
+Although cross-validation results in similar bias-variance reductions in both cases, in the case of linear regression $\hat{A}$ increases with increasing $k$ initially before leveling off.
+However, this is not surprising, as our linear regression simulation does not involve variable selection, so that the differences in models are more pronounced.
+To observe the cross-validation paradox in linear regression would require comparison of more similar linear models predisposed to similarity in $\widehat{\mathrm{PE}}$, as in the simulation of LASSO regression.
+
+In the cases of both linear and LASSO regression, $k_\mathrm{optimal}^*/n$ seems to change predictably with $n$.
+Specifically, it may be possible to model the relationship with some asymptotically decreasing function, while the relationship between $k_\mathrm{optimal}^*$ and $n$ seems to follow a positive pattern (Figures 9 and 10).
 
@@
 Instead, increasing $k$ results in bias and variance reduction.
-This phenomenon is observable in our simulation results, although it is also clear from the results that the reductions are not limitless, meaning that there is a point of diminishing returns in increasing $k$ and computational power allocation, as error estimation does not improve after a certain value of $k$.
+This phenomenon is observable in our simulation results, which suggest that bias and variance decrease asymptotically with increasing $k$.
 
-Our results also indicate a relationship between optimal fold number $k$ and sample size $n$, although more data is necessary to elucidate a trend.
-This is interesting as a future direction, as modeling the relationship between $n$ and $k_{optimal}$ would have practical utility, potentially improving the selection of $k$ from a largely arbitrary decision between 5 and 10.
+Our results also indicate a predictable relationship between $k_\mathrm{optimal}^*$ and sample size $n$.
+Although further data and analysis are needed to draw any reliable conclusions, modeling the relationship between $n$ and $k_\mathrm{optimal}^*$ would have practical utility, potentially improving the selection of $k$ from a largely arbitrary decision between 5 and 10.
+
 Future research may also focus on error estimation of other models, including models capturing non-linear relationships or involving tuning of multiple hyper-parameters, e.g., random forest or gradient boosting.
+It may also be interesting to study the value of $k$ in repeated cross-validation or nested cross-validation, with the value of $k$ variable in the inner loop, outer loop, or both.
+
+# Appendix
+
+```{r echo=FALSE, fig.height = 8, fig.cap = "Linear regression squared bias vs. fold number for various sample sizes. As fold number increases, bias decreases initially before leveling off."}
+source("~/optimal_k_git/plots/plots_linear.R")
+figure1
+```
+
+```{r echo=FALSE, fig.height = 8, fig.cap = "Linear regression variance vs. fold number for various sample sizes. As fold number increases, variance decreases initially before leveling off."}
+figure2
+```
+
+```{r echo=FALSE, fig.height = 8, fig.cap = "Linear regression mean squared error (MSE) vs.
+fold number for various sample sizes.
+As fold number increases, MSE decreases initially before leveling off."}
+figure3
+```
+
+```{r echo=FALSE, fig.cap = "LASSO regression squared bias vs. fold number for various sample sizes. As fold number increases, bias decreases initially before leveling off."}
+source("~/optimal_k_git/plots/plots_lasso.R")
+figure4
+```
+
+```{r echo=FALSE, fig.cap = "LASSO regression variance vs. fold number for various sample sizes. As fold number increases, variance decreases initially before leveling off."}
+figure5
+```
+
+```{r echo=FALSE, fig.cap = "LASSO regression mean squared error (MSE) vs. fold number for various sample sizes. As fold number increases, MSE decreases initially before leveling off."}
+figure6
+```
+
+```{r echo=FALSE, fig.height = 8, fig.cap = "Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
+figure7
+```
+
+```{r echo=FALSE, fig.cap = "Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
+figure8
+```
+
+```{r echo=FALSE, fig.cap = "Raw optimal fold number vs. sample size (top); fold number proportion of sample size vs. sample size (bottom)."}
+figure9
+```
 

### 2023-05-31 manuscript/manuscript.tex
“May31”

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
+\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t} \usepackage[labelsep=period]{caption}
 \usepackage{booktabs}
@@
 \hypersetup{
-  pdftitle={test},
+  pdftitle={Simulations of fold number selection in cross-validation of linear and LASSO regression},
   pdfauthor={Angelos Vasilopoulos; Gregory J. Matthews},
@@
 
-\title{test}
+\title{Simulations of fold number selection in cross-validation of
+linear and LASSO regression}
 \author{Angelos Vasilopoulos \and Gregory J. Matthews}
@@
 \begin{abstract}
-Abstract \vspace{2mm}\\
-\emph{Keywords}: Cross validation
+\emph{k}-fold cross-validation is a popular method of error estimation
+for model selection in computational research. However, there is limited
+focus in the literature on the question of what fold number \emph{k} is
+appropriate for various dataset dimensions. Here we review relevant
+literature and present a simulation of linear and least absolute
+shrinkage and selection operator (LASSO) regression prediction error
+estimation at various values of \emph{k} and sample size \emph{n}. In
+agreement with a growing body of literature, we find that contrary to a
+persisting understanding, there is no bias-variance trade-off in
+selection of \emph{k}. Instead, with increasing \emph{k} both bias and
+variance decrease, perhaps asymptotically. Our results also suggest a
+predictable relationship between optimal values of \emph{k} and
+\emph{n}. \textbar{} \emph{Keywords}: cross-validation, optimization,
+fold number
 \end{abstract}
-
-\newpage
 
@@
 \(k\)-fold cross-validation is a popular method of error estimation for
-model selection in computational research.
-
-n observations are divided into \(k\) groups. In a first iteration,
-\(i = k - 1\) groups are used as a training set. The remaining group is
-used as a test set to estimate prediction error. This process is
-repeated \(i\) times, with a different test set in each iteration. The
-average of \(k\) prediction errors (\(PE\))
-\[\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}PE_i\]
-
+model selection in computational research. \(n\) observations are
+divided into \(k\) groups. In a first iteration, \(i = k - 1\) groups
+are used as a training set. The remaining group is used as a test set to
+calculate estimated prediction error (\(\mathrm{\widehat{PE}}\)). This
+process is repeated \(i\) times, with a different test set in each
+iteration. The average of \(k\) estimates of \(\mathrm{PE}\)
+\[\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i\]
 is meant to estimate the true model error \(\theta\), i.e., the error of
@@
 the literature on the question of what fold number \(k\) is appropriate
-for cross-validation with a given data set. One popular idea is that
-selection of \(k\) comes with a bias-variance trade-off, specifically,
-that as \(k\) increases, the bias of k-fold error estimation
-\[Bias(\hat{\theta}) = E(\hat{\theta}) - \theta\] and variance of
-\(k\)-fold error estimation
-\[Var(\hat{\theta}) = E(\hat{\theta}^2) - E(\hat{\theta})^2\] as in
-\[Err(\hat{\theta}) = Bias^2(\hat{\theta}) + Var(\hat{\theta})\]
-decrease and increase, respectively. This idea appears in early
-literature (Efron (\protect\hyperlink{ref-Efron1983}{1983}); Kohavi
-(\protect\hyperlink{ref-Kohavi2001}{2001})).
+for cross-validation with a dataset of a given size. One popular idea is
+that selection of \(k\) comes with a bias-variance
+trade-off---specifically, that as \(k\) increases, the bias
+\(Bias(\hat{\theta}) = \mathrm{E}(\hat{\theta}) - \theta\) and variance
+\(Var(\hat{\theta}) = \mathrm{E}(\hat{\theta}^2) - \mathrm{E}(\hat{\theta})^2\)
+of \(k\)-fold error estimation decrease and increase, respectively. This
+idea appears in early literature (Efron
+(\protect\hyperlink{ref-Efron1983}{1983}); Kohavi
+(\protect\hyperlink{ref-Kohavi2001}{2001})) but also in modern, widely
+used textbooks (James et al. (\protect\hyperlink{ref-James2021}{2021})).
 
 Subsequent, albeit limited, literature argues differently. In the case
-of leave-one-out cross-validation, i.e., with \(k\) = 1, previous
-literature suggests that asymptotically both bias and variance of error
+of leave-one-out cross-validation (LOOCV), i.e., with \(k = n\), some
+authors suggest that asymptotically both bias and variance of error
 estimation decrease as \(k\) increases (Burman
@@
 More recently have been proposed ways to quantify the variance reduction
-achieved by cross-validation when the true prediction error is not
-known, e.g., as mean-square stability (Kale, Kumar, and Vassilvitskii
-(\protect\hyperlink{ref-Kale2011}{2011})), which measures performance
-estimate variance across folds, or as loss stability (Kumar et al.
-(\protect\hyperlink{ref-Kumar2013}{2013})), the proportion of times that
-a model outperforms competing models in all folds. However, it has also
-been demonstrated that, due to overlap between training and test sets in
+achieved by cross-validation when the true prediction error
+(\(\mathrm{PE}\)) is not known, e.g., as mean-square stability (Kale,
+Kumar, and Vassilvitskii (\protect\hyperlink{ref-Kale2011}{2011})),
+where
+\[\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,\]
+i.e., the expectation of the squared difference in loss \(\ell\) of an
+algorithm \(\mathcal A\) over a dataset \(S\) and over a dataset \(S'\)
+such that one element \(z' \sim Z\) has been added to \(S\) is less than
+\(\beta\), which depends on the algorithm; or as loss stability (Kumar
+et al. (\protect\hyperlink{ref-Kumar2013}{2013})), where
+\[\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,\]
+\(\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)\),
+and \(\gamma\) depends on the algorithm. However, it has also been
+demonstrated that, due to overlap between training and test sets in
 cross-validation, there is no universal (i.e., valid under all
-distributions) unbiased estimator of the variance of \(k\)-fold cross
-validation (Bengio and Grandvalet
+distributions) unbiased estimator of the variance of \(k\)-fold
+cross-validation (Bengio and Grandvalet
 (\protect\hyperlink{ref-Bengio2004}{2004})).
 
-In addition to theoretical analysis of variance reduction by cross
-validation, there are some simulation results showing this phenomenon
-(Zhang and Yang (\protect\hyperlink{ref-Zhang2015}{2015})). However,
-simulations currently in the literature provide limited insight into the
-dependence of optimal fold number \(k\) on sample size \(n\). Here we
-present a simulation of linear regression and least absolute shrinkage
-and regression operator (LASSO) regression to observe the relationship
-of cross-validation fold number \(k\) and model prediction error
-estimation accuracy for various sample sizes \(n\).
+In addition to theoretical analyses of variance reduction by
+cross-validation, there are some simulation results showing this
+phenomenon. However, simulations currently in the literature provide
+limited insight into the dependence of optimal fold number
+\(k_\mathrm{optimal}\) on sample size \(n\) (Zhang and Yang
+(\protect\hyperlink{ref-Zhang2015}{2015})) or involve biased variance
+calculations (Marcot and Hanea
+(\protect\hyperlink{ref-Marcot2021}{2021})). Here we present a
+simulation of linear regression and least absolute shrinkage and
+selection operator (LASSO) regression to observe the relationship of
+cross-validation fold number \(k\) to model selection accuracy for
+various samples of size \(n\) of a known population.
 
@@
 
-Consider a population of size \(N = 500,000\) with five of 100 features
-\(X \sim N(0, 1)\) a linear combination
-\[Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5 + \epsilon\]
-where \(\beta_1...\beta_5 = 1\) and \(\epsilon \sim N(0, 10)\). Then the
-true model of the population is
-\[E(Y) = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4X_4 + \beta_5X_5\]
-and the mean squared error (\(MSE\)) of the true model is
-\[MSE[E(Y)] = \theta = \frac{\sum_{i=1}^{N}[Y - E(Y)]^2}{N}.\]
+Consider a population of size N = 500,000 with five of \(P = 100\)
+features \(X_1...X_{100} \sim N(0, 1)\) a linear combination of feature
+\(Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon\) where
+\(\beta_1 = \cdots = \beta_5 = 1\) and \(\epsilon \sim N(0, 10)\). The
+true model of \(Y\) is \(f \in F\), a set of competing models.
+\(f = \mathrm{E}(Y) = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5\) and
+its mean squared error (\(\mathrm{MSE}\)) is
+\[\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.\]
 
 From the population we take a sample of size \(n\). We estimate \(Y\) as
-\(\hat{Y}\) by regression on a subset of \(X\) and estimate
-\(MSE(\hat{Y})\) as
-\[\hat{\theta} = \frac{\sum_{i=1}^{N}(Y - \hat{Y})^2}{N}\] by \(k\)-fold
-cross-validation. We repeat this procedure in 1000 simulations for the
-same subsets of \(X\), including the five-variable subset of the true
-model, and values of \(k\). We then calculate simulation-wise
-\(MSE(\hat{\theta})\) for each value of \(k\).
+\(\hat{Y}\) by regression on a subset of \(X_1...X_{100}\), regression
+coefficients estimated by the least squares method, and compute
+\(\mathrm{MSE}(\hat{Y})\) by \(k\)-fold cross-validation as
+\[\hat{\theta} = \mathrm{MSE}(\hat{Y}) = \frac{1}{k}\sum_{j=1}^{k}\frac{\sum_{i=1}^{n/k}(Y_{j_i} - \hat{Y_{j_i}})^2}{n/k}\]
+where \(j_i\) is the index of the \(i^{th}\) element of the \(j^{th}\)
+fold.
+
+In addition to regression with \(X_1...X_5\), \(f\), we under-fit with
+\(X_1...X_3\), over-fit with \(X_1...X_{20}\) and \(X_1...X_{100}\), and
+do regression with noise \(X_6...X_{100}\) only, for each \(k \in\) \{2,
+10, 20, 30, \ldots, \(n\)\} for each \(n \in\) \{100, 200, 300, \ldots,
+1000\}. We perform this simulation 1000 times and then calculate
+simulation-wise \(\mathrm{MSE}(\hat{\theta})\) for each \(k\), for each
+\(n\).
 
 We perform an additional 1000 simulations predicting \(Y\) as
-\(\hat{Y}\) by LASSO regression instead of linear regression. To avoid
-data leakage, we introduce an inner 5-fold cross-validation loop for the
-selection of regularization parameter \(\lambda\) as in the minimization
-of
-\[\sum_{i=1}^{n}(Y_i - \sum_{j}X_{ij}\beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.\]
+\(\hat{Y}\) by LASSO regression instead of linear regression for each
+\(k \in\) \{2, 10, \(n\)/10, 2\(n\)/10, 3\(n\)/10, \ldots, \(n\)\} for
+each \(n \in\) \{250, 500, 750, 1000\}. To avoid data leakage, we
+introduce an inner 5-fold cross-validation loop for the selection of
+parameter \(\lambda\) as in the minimization of
+\[\sum_{i=j}^{p}(Y_j - \beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.\]
 
-We perform the same 1000 simulations with \(n\) = \{100, 500, 1000\} and
-consider as the optimal value of \(k\) the value \(k_{optimal}\)
-resulting in the lowest \(MSE(\hat{\theta})\) for the true model the
-greatest number of times.
+For each \(k\), for each \(n\), we count the number of times \(A\) that
+the true model is selected and consider as optimal fold number for each
+\(n\) the value \(k_\mathrm{optimal}\), at which \(A\) is the highest.
 
-This is because in practice the model with the lowest prediction error
-is more likely to be selected. We refer to this as lowest-error model
-selection. As our simulation results show, however, it is possible for
-\(k\)-fold cross validation to result in calculations of
-\(MSE(\hat{\theta})\) that suggest that a competing model has a lower
-prediction error than the true model. This false model may have the
-lowest prediction error in \(k\)-fold cross validation but its
-generalization error will be lower in the long-run (i.e., when tested on
-a large part of the population) than the generalization error of the
-true model, as \(E(Y)\) is a linear combination of the true model only.
-Thus, it is preferable for the value of \(k\) selected to result in the
-true model having the lowest estimated prediction error.
+This is because in practice the model with the lowest
+\(\mathrm{\widehat{PE}}\) is more likely to be selected. We refer to
+this as lowest-error model selection. As our simulation results show,
+however, it is possible for \(k\)-fold cross-validation to result in
+calculations of \(\mathrm{MSE}(\hat{\theta})\) such that a competing
+model \(\hat{f}\) has a lower \(\mathrm{\widehat{PE}}\) than the true
+model \(f\). This false model may have the lowest
+\(\mathrm{\widehat{PE}}\) but its generalization error will be higher in
+the long-run (i.e., when tested on a large part of the population) than
+the generalization error of the true model. Thus, it is preferable for
+the value of \(k\) selected to result in \(f\) having the lowest
+\(\mathrm{\widehat{PE}}\).
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+Having considered the effect of \(k\) on model selection accuracy for a
+given \(n\), we may also consider the relationship of
+\(k_\mathrm{optimal}\) and \(n\). Our results suggest that after a
+certain value of \(k\), changes in \(A\) become negligible, in the case
+of linear regression, or negative, in the case of LASSO regression, and
+resource-expensive increases of fold number become undesirable. We refer
+to this ``point of diminishing returns'' as \(k_\mathrm{optimal}^*\).
 
-We find that for all \(n\) as fold number \(k\) increases, both bias and
-variance of error estimation decrease initially before leveling off
-(Figures 1 - 3).
+To estimate \(k_\mathrm{optimal}*\) for each \(n\), we first fit to the
+dataset \((k,A)\) for each sample size \(n\) a non-linear least squares
+model of the form \[\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}\] with
+constants \(B\) to \(E\) specific to each \(n\) and constants \(B\) and
+\(E\) set to zero for linear regression. We then draw a line \(L\)
+through \((k_1, \hat{A}_1)\) and \((k_n, \hat{A}_n)\) and maximize the
+perpendicular distance \(d\) of each point on \(\hat{A}\) from \(L\), so
+that \(k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}\).
 
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\hypertarget{results-and-discussion}{%
+\section{Results and discussion}\label{results-and-discussion}}
 
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-2} \end{center}
+Interpretations of early literature have resulted in lasting
+misconceptions about the use of cross-validation. Such misconceptions
+include the idea that there is a bias-variance trade-off
+\(Bias^2(\hat{\theta}) \propto 1/Var(\hat{\theta})\) associated with
+selection of \(k\) and that \(k = 10\) is the best value to use in
+\(k\)-fold cross-validation (Kohavi
+(\protect\hyperlink{ref-Kohavi2001}{2001})).
 
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-3} \end{center}
+In agreement with a small but growing body of literature, our simulation
+results suggest that neither of these ideas are necessarily correct.
+Instead, in the context of linear and LASSO regression with standard
+normal data and certain random error, we find that for various \(n\)
+both bias and variance decrease as \(k\) increases (Figures 1 - 6),
+i.e., \(Bias^2(\hat{\theta}) \propto Var(\hat{\theta})\), and although
+in the case of LASSO, 10-fold cross-validation seems to be a
+near-optimal choice for large \(n\), for smaller samples and linear
+regression other values of \(k\) appear to be optimal for model
+selection (Figures 7 and 8).
 
-We also find that for larger \(n\), e.g., \(n = 1000\), lowest-error
-model selection becomes less reliable as \(k\) increases past a certain
-value, i.e., as \(k\) increases past a certain value the true model has
-the lowest MSE less frequently. With increasing \(n\) we also observe
-variability in the improvement of lowest-error model selection
-reliability (Figure 4; Table 1).
+As previous authors have noted (Zhang and Yang
+(\protect\hyperlink{ref-Zhang2015}{2015})), it is important to
+distinguish between possible goals of cross-validation that previously
+have been conflated (Kohavi (\protect\hyperlink{ref-Kohavi2001}{2001})):
+estimation of \(\mathrm{PE}\), in which case
+\(k_\mathrm{optimal} = {\arg\min}_{k}(\mathrm{PE} - \mathrm{\widehat{PE}})\),
+and model selection, in which case
+\[k_\mathrm{optimal} = {\arg\max}_k(A)\], which is related to the
+definition of \(k_\mathrm{optimal}^*\) used in this study.
 
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1} \end{center}
-\begin{table}[!h]
-\centering
-\begin{tabular}{r|r|r|r|r}
-\hline
-n & Minimum & Maximum & Range & Variance\\
-\hline
-100 & 240 & 306 & 66 & 140.0310\\
-\hline
-250 & 386 & 500 & 114 & 427.1585\\
-\hline
-500 & 619 & 813 & 194 & 676.5341\\
-\hline
-1000 & 861 & 931 & 70 & 711.2000\\
-\hline
-\end{tabular}
-\end{table}
+We find that in the cases of both linear and LASSO regression, as \(n\)
+increases, \(k_\mathrm{optimal}^*\) increases but at a lower rate than
+\(n\), such that \(k_\mathrm{optimal}^*/n\) decreases, perhaps
+asymptotically. For LASSO regression, we also find that as \(k\)
+increases past a certain value, lowest-error model selection becomes
+less reliable, i.e., \(\hat{A}\) decreases, i.e., the true model has the
+lowest \(\mathrm{MSE}\) less frequently (Figure 8). The rate of this
+reduction increases with \(n\).
+
+The reason for this is related to what is known as the cross-validation
+paradox (Yang (\protect\hyperlink{ref-Yang2006}{2006})): greater
+quantity of data results in more accurate estimation of \(\mathrm{PE}\);
+occasionally, this makes model-wise differences in
+\(\mathrm{\widehat{PE}} = CV_{(k)}\) less exaggerated, making it more
+difficult to distinguish between models and resulting in less accurate
+model selection.
+
+However, we do not observe the cross-validation paradox in our linear
+regression simulations. Although cross-validation results in similar
+bias-variance reductions in both cases, in the case of linear regression
+\(\hat{A}\) increases with increasing \(k\) initially before leveling
+off. However, this is not surprising, as our linear regression
+simulation does not involve variable selection, so that the differences
+in models are more pronounced. To observe the cross-validation paradox
+in linear regression would require comparison of more similar linear
+models predisposed to similarity in \(\widehat{\mathrm{PE}}\), as in the
+simulation of LASSO regression.
+
+In the cases of both linear and LASSO regression,
+\(k_\mathrm{optimal}^*/n\) seems to change predictably with \(n\).
+Specifically, it may be possible to model the relationship with some
+asymptotically decreasing function, while the relationship between
+\(k_\mathrm{optimal}^*\) and \(n\) seems to follow a positive pattern
+(Figures 9 and 10).
 
@@
 case. Instead, increasing \(k\) results in bias and variance reduction.
-This phenomenon is observable in our simulation results, although it is
-also clear from the results that the reductions are not limitless,
-meaning that there is a point of diminishing returns in increasing \(k\)
-and computational power allocation, as error estimation does not improve
-after a certain value of \(k\).
+This phenomenon is observable in our simulation results, which suggest
+that bias and variance decrease asymptotically with increasing \(k\).
 
-Our results also indicate a relationship between optimal fold number
-\(k\) and sample size \(n\), although more data is necessary to
-elucidate a trend. This is interesting as a future direction, as
-modeling the relationship between \(n\) and \(k_{optimal}\) would have
+Our results also indicate a predictable relationship between
+\(k_\mathrm{optimal}^*\) and sample size \(n\). Although further data
+and analysis are needed to draw any reliable conclusions, modeling the
+relationship between \(n\) and \(k_\mathrm{optimal}^*\) would have
 practical utility, potentially improving the selection of \(k\) from a
-largely arbitrary decision between 5 and 10. Future research may also
-focus on error estimation of other models, including models capturing
-non-linear relationships or involving tuning of multiple
-hyper-parameters, e.g., random forest or gradient boosting.
+largely arbitrary decision between 5 and 10.
+
+Future research may also focus on error estimation of other models,
+including models capturing non-linear relationships or involving tuning
+of multiple hyper-parameters, e.g., random forest or gradient boosting.
+It may also be interesting to study the value of \(k\) in repeated
+cross-validation or nested cross-validation, with the value of \(k\)
+variable in the inner loop, outer loop, or both.
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1}
+
+}
+
+\caption{Linear regression squared bias vs. fold number for various sample sizes. As fold number increases, bias decreases initially before leveling off.}\label{fig:unnamed-chunk-1}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1}
+
+}
+
+\caption{Linear regression variance vs. fold number for various sample sizes. As fold number increases, variance decreases initially before leveling off.}\label{fig:unnamed-chunk-2}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1}
+
+}
+
+\caption{Linear regression mean squared error (MSE) vs. fold number for various sample sizes. As fold number increases, MSE decreases initially before leveling off.}\label{fig:unnamed-chunk-3}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1}
+
+}
+
+\caption{LASSO regression squared bias vs. fold number for various sample sizes. As fold number increases, bias decreases initially before leveling off.}\label{fig:unnamed-chunk-4}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1}
+
+}
+
+\caption{LASSO regression variance vs. fold number for various sample sizes. As fold number increases, variance decreases initially before leveling off.}\label{fig:unnamed-chunk-5}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{LASSO regression mean squared error (MSE) vs. fold number for various sample sizes. As fold number increases, MSE decreases initially before leveling off.}\label{fig:unnamed-chunk-6}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1}
+
+}
+
+\caption{Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-7}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1}
+
+}
+
+\caption{Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-8}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-9-1}
+
+}
+
+\caption{Raw optimal fold number vs. sample size (top); fold number proportion of sample size vs. sample size (bottom).}\label{fig:unnamed-chunk-9}
+\end{figure}
 
@@
 Bengio, Yoshua, and Yves Grandvalet. 2004. {``No Unbiased Estimator of
-the Variance of k-Fold Cross-Validation,''} June.
+the Variance of k-Fold Cross-Validation.''} \emph{Journal of Machine
+Learning Research} 5 (June): 1089--1105.
 
@@
 Evaluation in Regression. The x-Random Case.''} \emph{International
-Statistical Review / Revue Internationale de Statistique} 60 (3): 291.
-\url{https://doi.org/10.2307/1403680}.
+Statistical Review / Revue Internationale de Statistique} 60 (December):
+291.
 
@@
 Learning-Testing Methods.''} \emph{Biometrika} 76 (September): 503--14.
-\url{https://doi.org/10.1093/biomet/76.3.503}.
 
@@
 Improvement on Cross-Validation.''} \emph{Journal of the American
-Statistical Association} 78 (382): 316--31.
-\url{https://doi.org/10.1080/01621459.1983.10477973}.
+Statistical Association} 78 (March): 316--31.
+
+\leavevmode\vadjust pre{\hypertarget{ref-James2021}{}}%
+James, Gareth, Daniela Witten, Trevor Hastie, and Robert Tibshirani.
+2021. \emph{An Introduction to Statistical Learning}. 2nd ed. Springer.
 
@@
 Kale, Satyen, Ravi Kumar, and Sergei Vassilvitskii. 2011.
-{``Cross-Validation and Mean-Square Stability.''} In, 487--95.
+{``Cross-Validation and Mean-Square Stability.''} In \emph{International
+Conference on Supercomputing}, 487--95.
 
@@
 Kohavi, Ron. 2001. {``A Study of Cross-Validation and Bootstrap for
-Accuracy Estimation and Model Selection''} 14 (March).
+Accuracy Estimation and Model Selection.''} \emph{International Joint
+Conference on Artificial Intelligence} 14 (March).
 
@@
 Stability.''} In \emph{Proceedings of the 30th International Conference
-on Machine Learning}, edited by Sanjoy Dasgupta and David McAllester,
-28:27--35. Proceedings of Machine Learning Research 1. Atlanta, Georgia,
-USA: PMLR. \url{https://proceedings.mlr.press/v28/kumar13a.html}.
+on Machine Learning}, 28:27--35. Proceedings of Machine Learning
+Research. PMLR.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Marcot2021}{}}%
+Marcot, Bruce, and Anca Hanea. 2021. {``What Is an Optimal Value of k in
+k-Fold Cross-Validation in Discrete Bayesian Network Analysis?''}
+\emph{Computational Statistics} 36 (September): 2009--31.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Yang2006}{}}%
+Yang, Yuhong. 2006. {``Comparing Learning Methods for Classification.''}
+\emph{Statistica Sinica} 16 (April): 635--57.
 
@@
 Zhang, Yongli, and Yuhong Yang. 2015. {``Cross-Validation for Selecting
-a Model Selection Procedure.''} \emph{Journal of Econometrics} 187 (1):
-95--112. \url{https://doi.org/10.1016/j.jeconom.2015.02.006}.
+a Model Selection Procedure.''} \emph{Journal of Econometrics} 187
+(February): 95--112.
 

### 2023-05-31 plots/plots_lasso.R

@@ -38,12 +38,12 @@ k <- list(n100 = seq(2, 100, 2),
           n500 = c(2, seq(10, 500, 10)),
           n1000 = c(2, 10, 100, 500, 1000)) # for n = 1000
 n_mod <- length(variables)
-out <- list(n100 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
-            n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
-            n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
-            n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
+out <- list(n100 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
+            n250 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
+            n500 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
+            n1000 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
 titles <- list("n = 100", "n = 250", "n = 500", "n = 1000")
-plotsout <- list(list(), list(), list())
+plotsout <- list(list(), list(), list(), list())
 figures <- list()
 
 for(i in 1:length(out)){
@@ -96,16 +96,16 @@ for(i in 1:length(metrics)){
 
 
 
-figure1 <- annotate_figure(figures[[1]],
-                top = text_grob("Squared bias vs. k", size = 10, x = 0.4),
+figure4 <- annotate_figure(figures[[1]],
+                top = text_grob("LASSO regression: squared bias vs. k", size = 10, x = 0.4),
                 bottom = text_grob("k", size = 10, x = 0.4),
                 left = text_grob("Squared bias", size = 10, rot = 90))
-figure2 <- annotate_figure(figures[[2]],
-                top = text_grob("Variance vs. k", size = 10, x = 0.4),
+figure5 <- annotate_figure(figures[[2]],
+                top = text_grob("LASSO regression: variance vs. k", size = 10, x = 0.4),
[…]

### 2023-05-31 plots/plots_linear.R

@@ -1,0 +1,244 @@
+
+# load libraries
+library(ggplot2)
+library(ggpubr)
+library(kableExtra)
+
+# create population
+N <- 500000
+x<-data.frame(matrix(nrow=500000,ncol=100))
+set.seed(2022)
+n_var <- 1:100
+for(i in n_var){
+  x[,i]<-rnorm(n=N,mean=0,sd=1)
+}
+
+# define parameters
+beta0 <- 0
+beta1 <- 1
+beta2 <- 1
+beta3 <- 1
+beta4 <- 1
+beta5 <- 1
+sigma <- 10
+
+# define theta
+Ey <- beta0 + beta1*x[,1] + beta2*x[,2] + beta3*x[,3] + beta4*x[,4] + beta5*x[,5]
+y <- Ey + rnorm(nrow(x),0,sigma)
+theta <- mean((y - Ey)^2)
+
+# choose models
+variables <- list(1:3, 1:5, 1:20, 1:100, 6:100)
+
+# loop settings
+n <- seq(100, 1000, 100)
+nsim <- 1000
+k <- list(n100 = c(2, seq(10, 100, 10)),
+          n200 = c(2, seq(10, 200, 10)),
+          n300 = c(2, seq(10, 300, 10)),
+          n400 = c(2, seq(10, 400, 10)),
+          n500 = c(2, seq(10, 500, 10)),
+          n600 = c(2, seq(10, 600, 10)),
+          n700 = c(2, seq(10, 700, 10)),
+          n800 = c(2, seq(10, 800, 10)),
+          n900 = c(2, seq(10, 900, 10)),
+          n1000 = c(2, seq(10, 1000, 10)))
+n_mod <- length(variables)
+out <- list(n100 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_100_n_100.rds"),
+            n200 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_200_n_200.rds"),
+            n300 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_300_n_300.rds"),
+            n400 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_400_n_400.rds"),
+            n500 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_500_n_500.rds"),
[…]

### 2023-05-31 simulations/simulation_lasso_parallel.R

@@ -1,3 +1,4 @@
+
 # create population
 N <- 500000
 x<-data.frame(matrix(nrow=500000,ncol=100))
@@ -22,17 +23,13 @@ y <- Ey + rnorm(nrow(x),0,sigma)
 theta <- mean((y - Ey)^2)
 
 # choose models
-variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
+variables <- list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- 750 # 250 # 500 # 100 # 1000
+# 2, 10, 20, 30, ..., 100, n/2, n
+n <- 1000
+k <- c(2, 10, seq(n/10, n, n/10))
 nsim <- 1000
-# k <- c(seq(2, 100, 2), 500, 1000)
-# k <- seq(2, 100, 2) # for n = 100 and n = 1000
-# k <- c(2, seq(10, 250, 10)) # for n = 250
-k <- c(2, seq(10, 500, 10)) # for n = 500
-# k <- c(2, seq(10, 750, 10)) # for n = 750
-# k <- c(2, 10, 100, 500, 1000) # for n = 1000
 n_mod <- length(variables)
 
 # create objects
@@ -71,7 +68,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("~/optimal_k_git/results/parallel_simulation_lasso_output_k_2_500_n_500 copy.rds")
+out <- readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_100_n_1000.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -115,12 +112,25 @@ for(sim in 1:nsim){
   results[sim, ] <- apply(theta.hat[,, sim], 1, which.min)
 }
 corrects <- colSums(results == 2)
-plot(k, corrects, type = "b", ylab = "How many times correct model had lowest MSE (n = 250)")
-df <- data.frame(k, corrects)
+plot(k, corrects)
 
-# plot
-samplesize <- c(100, 250, 500, 1000)
-optimalk <- c(24, 80, 70, 4)
-abline(plot(samplesize, optimalk))
+# nls
[…]

### 2023-05-31 simulations/simulation_linear_parallel.R

@@ -26,7 +26,7 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- 500 # 1000
+n <- 400
 nsim <- 1000
 k <- c(2, seq(10, n, 10)) # c(2, 10, 100, 500, 1000)
 n_mod <- length(variables)
@@ -65,7 +65,7 @@ out <- foreach(j = 1:nsim, .combine = "list") %dopar% {
 }
 
 # read results
-out <- readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_500_n_500.rds")
+out <- readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_400_n_00.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim), dimnames = list(rep("k", length(k)), rep("mod", n_mod), rep("sim", nsim)))
 
 # biassq + var = mse
@@ -109,37 +109,50 @@ for(sim in 1:nsim){
   results[sim, ] <- apply(theta.hat[,, sim], 1, which.min)
 }
 corrects <- colSums(results == 2)
-df <- data.frame(k, corrects)
-plot(df$k, df$corrects)
+plot(k, corrects)
 
 # nls
-nls.fit <- nls(corrects ~ a*k^-1 + b, start = list(a = 14.5, b = 660))
-.a <- coef(nls.fit)[1]
-.b <- coef(nls.fit)[2]
-eq <- function(x){.a*(x)^-1 + .b}
-lines(eq(0:700), type = "l")
+nls.fit <- nls(corrects ~ c + d*k^-1, start = list(c = 970, d = -208))
+.c <- coef(nls.fit)[1]
+.d <- coef(nls.fit)[2]
+eq <- function(x){.c + .d*x^-1}
+pred <- eq(2:n)
+lines(pred, type = "l")
+
+# elbow point, optimal.k/n vs. n
+library("smerc")
+elbow_point(2:n, pred)$x
+samp.n <- seq(100, 1000, 100)
+opt.k <- c(14/100, 20/200, 24/300, 28/400, 32/500, 35/600, 37/700, 40/800, 42/900, 45/1000) # departing from LOOCV
+plot(samp.n, opt.k)
+
+
+
+
+
+
[…]

### 2023-06-02 manuscript/manuscript.Rmd
“June2”

@@
 
-We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression for each $k \in$ {2, 10, $n$/10, 2$n$/10, 3$n$/10, ..., $n$} for each $n \in$ {250, 500, 750, 1000}.
+We perform an additional 1000 simulations predicting $Y$ as $\hat{Y}$ by LASSO regression instead of linear regression for each $k \in$ {2, 10, 20, 30, ..., 100, $n$} for each $n \in$ {250, 500, 750, 1000}.
 To avoid data leakage, we introduce an inner 5-fold cross-validation loop for the selection of parameter $\lambda$ as in the minimization of
@@
 with constants $B$ to $E$ specific to each $n$ and constants $B$ and $E$ set to zero for linear regression.
-We then draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that $k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}$.
+We then draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that
+$$k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.$$
 

### 2023-06-02 manuscript/manuscript.tex
“June2”

@@
 \(\hat{Y}\) by LASSO regression instead of linear regression for each
-\(k \in\) \{2, 10, \(n\)/10, 2\(n\)/10, 3\(n\)/10, \ldots, \(n\)\} for
-each \(n \in\) \{250, 500, 750, 1000\}. To avoid data leakage, we
-introduce an inner 5-fold cross-validation loop for the selection of
-parameter \(\lambda\) as in the minimization of
+\(k \in\) \{2, 10, 20, 30, \ldots, 100, \(n\)\} for each \(n \in\)
+\{250, 500, 750, 1000\}. To avoid data leakage, we introduce an inner
+5-fold cross-validation loop for the selection of parameter \(\lambda\)
+as in the minimization of
 \[\sum_{i=j}^{p}(Y_j - \beta_j)^2 + \lambda\sum_{j=1}^{p}|\beta_j|.\]
@@
 perpendicular distance \(d\) of each point on \(\hat{A}\) from \(L\), so
-that \(k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}\).
+that \[k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.\]
 

### 2023-06-02 plots/plots_lasso.R

@@ -31,18 +31,18 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- c(100, 250, 500, 1000)
+n <- c(250, 500, 750, 1000)
 nsim <- 1000
-k <- list(n100 = seq(2, 100, 2),
-          n250 = c(2, seq(10, 250, 10)),
-          n500 = c(2, seq(10, 500, 10)),
-          n1000 = c(2, 10, 100, 500, 1000)) # for n = 1000
+k <- list(n250 = c(2, seq(10, 100, 10), n[1]),
+          n500 = c(2, seq(10, 100, 10), n[2]),
+          n750 = c(2, seq(10, 100, 10), n[3]),
+          n1000 = c(2, seq(10, 100, 10), n[4])) # for n = 1000
 n_mod <- length(variables)
-out <- list(n100 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
-            n250 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
-            n500 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
-            n1000 = readRDS("~/optimal_k_git/results/extra/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
-titles <- list("n = 100", "n = 250", "n = 500", "n = 1000")
+out <- list(n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_250.rds"),
+            n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_500.rds"),
+            n750 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_750.rds"),
+            n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_1000.rds"))
+titles <- list("n = 250", "n = 500", "n = 750", "n = 1000")
 plotsout <- list(list(), list(), list(), list())
 figures <- list()
 
@@ -93,8 +93,6 @@ for(i in 1:length(metrics)){
                             align = "hv", 
[…]

### 2023-06-02 simulations/simulation_lasso_parallel.R

@@ -27,8 +27,8 @@ variables <- list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
 # 2, 10, 20, 30, ..., 100, n/2, n
-n <- 1000
-k <- c(2, 10, seq(n/10, n, n/10))
+n <- 750
+k <- c(2, seq(10, n, 10))
 nsim <- 1000
 n_mod <- length(variables)
 
@@ -68,7 +68,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_100_n_1000.rds")
+out <- readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_1000.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -124,13 +124,9 @@ eq <- function(x){.a*x + .b + .c*x^-1 + .d*x^-2}
 pred <- eq(2:n)
 lines(pred, type = "l")
 
-# elbow point, optimal.k/n vs. n (n/10 scheme)
+# elbow point
 library("smerc")
 elbow_point(2:n, pred)$x
 samp.n <- c(250, 500, 750, 1000)
-opt.k <- c(17, 17, 8, 8)
+opt.k <- c(17, 17, 8, 5)
 plot(samp.n, opt.k/samp.n)
-
-# elbow point, optimal.k/n vs. n (n/10 scheme)
-
-

### 2023-07-20 manuscript/manuscript.Rmd
“July20”

@@
 $$\hat{\theta} = CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k}\mathrm{\widehat{PE}}_i$$
-is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population.
+is meant to estimate the true model error $\theta$, i.e., the error of the model tested on the population (@Bates2022).
 
@@
 
-More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011), where
-$$\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,$$
-i.e., the expectation of the squared difference in loss $\ell$ of an algorithm $\mathcal A$ over a dataset $S$ and over a dataset $S'$ such that one element $z' \sim Z$ has been added to $S$ is less than $\beta$, which depends on the algorithm; or as loss stability (@Kumar2013), where
-$$\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,$$
-$\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)$, and $\gamma$ depends on the algorithm.
+More recently have been proposed ways to quantify the variance reduction achieved by cross-validation when the true prediction error ($\mathrm{PE}$) is not known, e.g., as mean-square stability (@Kale2011) or as loss stability (@Kumar2013).
 However, it has also been demonstrated that, due to overlap between training and test sets in cross-validation, there is no universal (i.e., valid under all distributions) unbiased estimator of the variance of $k$-fold cross-validation (@Bengio2004).
@@
 
+In machine learning tasks, selection of $k$ for $k$-fold cross validation is largely an arbitrary decision between, e.g., 5 and 10.
+With the results of the following simulation, we make a less arbitrary recommendation of $k$ with the aim of improving model selection accuracy.
+
 Consider a population of size N = 500,000 with five of $P = 100$ features $X_1...X_{100} \sim N(0, 1)$ a linear combination of feature $Y = \beta_0 + \beta_1X_1 + \cdots + \beta_5X_5 + \epsilon$
@@
 and its mean squared error ($\mathrm{MSE}$) is
-$$\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$
+$$\theta = \mathrm{MSE}(f) = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.$$
 
@@
 We refer to this as lowest-error model selection.
-As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model $\hat{f}$ has a lower $\mathrm{\widehat{PE}}$ than the true model $f$.
+As our simulation results show, however, it is possible for $k$-fold cross-validation to result in calculations of $\mathrm{MSE}(\hat{\theta})$ such that a competing model $\hat{f} \in F$ has a lower $\mathrm{\widehat{PE}}$ than the true model $f$.
 This false model may have the lowest $\mathrm{\widehat{PE}}$ but its generalization error will be higher in the long-run (i.e., when tested on a large part of the population) than the generalization error of the true model.
@@
 
-To estimate $k_\mathrm{optimal}*$ for each $n$, we first fit to the dataset $(k,A)$ for each sample size $n$ a non-linear least squares model of the form
-$$\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}$$
-with constants $B$ to $E$ specific to each $n$ and constants $B$ and $E$ set to zero for linear regression.
-We then draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that
-$$k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.$$
+To estimate $k_\mathrm{optimal}*$ for each $n$, we first fit to the dataset $(k,A)$ for each sample size $n$ a model of the form
+$$\hat{A}(k) = Bk + C + \frac{D}{k}$$
+with constants $B$ to $D$ specific to each $n$.
+Certain more complicated models (e.g., with higher-order terms) may provide better approximation of $A$ but make optimization by the following scheme non-trivial.
+We draw a line $L$ through $(k_1, \hat{A}_1)$ and $(k_n, \hat{A}_n)$ and maximize the perpendicular distance $d$ of each point on $\hat{A}$ from $L$, so that
+\begin{align*}
+k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[\hat{A}(k),L]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=m(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=\frac{(Bn+C+\frac{D}{n})-(2B+C+\frac{D}{2})}{n-2}(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=(B-\frac{D}{2n})(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]}. \\
+\end{align*}
+It is known that the distance $d$ between a point $(k, \hat{A}(k))$ and a line of the form
+$$ax + by + c = 0$$
+is
+\begin{align*}
+d = \frac{|ak + b\hat{A}(k) + c|}{\sqrt{a^2+b^2}}
+\end{align*}
+so
+\begin{align*}
+k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]} \\
+                       &= {\arg\max}_k{\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}}.
+\end{align*}
+Maximizing $d$, knowing that $n$ and $k$ are always positive, we find that
+\begin{align*}
+\frac{d}{dk}d[\hat{A}(k),L] &= \frac{d}{dk}\left(\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}\right) \\
+                            &=\frac{|n||k|(2nD-Dk^2)(-Dk^2-2nD+nDk+2Dk)}{nk^3|-Dk^2-2nD+nDk+2Dk|\sqrt{D^2+4n^2}} = 0 \\
+                            &\Rightarrow k = 0, 2, \sqrt{2n}, n.
+\end{align*}
+Evaluating $d$ at these four values of $k$, we find that
+\begin{align*}
+d(0) &= \mathrm{DNE} \\
+d(2) &= 0 \\
+d(\sqrt{2n}) &=  \frac{|nD+2D-\sqrt{2n}D-D|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}} \\
+d(n) &= 0.
+\end{align*}
+Since $d(\sqrt{2n}) > 0$,
+$$k^*_{\mathrm{optimal}} = \sqrt{2n}$$
+which fits the relationship of $k^*_{\mathrm{optimal}}$ vs. $n$ (Figure 9).
 
@@
 
-```{r echo=FALSE, fig.height = 8, fig.cap = "Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
+```{r echo=FALSE, fig.height = 8, fig.cap = "Count of true linear regression model selection by the lowest-error criterion vs.
+fold number for various sample sizes.
+Red line indicates optimal value of k."}
 figure7
@@
 
-```{r echo=FALSE, fig.cap = "Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes."}
+```{r echo=FALSE, fig.cap = "Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k."}
 figure8

### 2023-07-20 manuscript/manuscript.tex
“July20”

@@
 is meant to estimate the true model error \(\theta\), i.e., the error of
-the model tested on the population.
+the model tested on the population (Bates, Hastie, and Tibshirani
+(\protect\hyperlink{ref-Bates2022}{2022})).
 
@@
 (\(\mathrm{PE}\)) is not known, e.g., as mean-square stability (Kale,
-Kumar, and Vassilvitskii (\protect\hyperlink{ref-Kale2011}{2011})),
-where
-\[\underset{S,z,z'}{\mathrm{E}}[(\ell_{\mathcal A(S)}(z) - \ell_{\mathcal A(S^{i,z'})}(z))^2] \le \beta,\]
-i.e., the expectation of the squared difference in loss \(\ell\) of an
-algorithm \(\mathcal A\) over a dataset \(S\) and over a dataset \(S'\)
-such that one element \(z' \sim Z\) has been added to \(S\) is less than
-\(\beta\), which depends on the algorithm; or as loss stability (Kumar
-et al. (\protect\hyperlink{ref-Kumar2013}{2013})), where
-\[\underset{S,z,z'}{\mathrm{E}}[(\ell'_{\mathcal A(S)}(z) - \ell'_{\mathcal A(S^{i,z'})}(z))^2] \le \gamma,\]
-\(\ell'_{\mathcal A(S)}(z) = \ell_{\mathcal A(S)}(z) - \bar{\ell}_{\mathcal A(S)}(z)\),
-and \(\gamma\) depends on the algorithm. However, it has also been
-demonstrated that, due to overlap between training and test sets in
-cross-validation, there is no universal (i.e., valid under all
-distributions) unbiased estimator of the variance of \(k\)-fold
-cross-validation (Bengio and Grandvalet
+Kumar, and Vassilvitskii (\protect\hyperlink{ref-Kale2011}{2011})) or as
+loss stability (Kumar et al. (\protect\hyperlink{ref-Kumar2013}{2013})).
+However, it has also been demonstrated that, due to overlap between
+training and test sets in cross-validation, there is no universal (i.e.,
+valid under all distributions) unbiased estimator of the variance of
+\(k\)-fold cross-validation (Bengio and Grandvalet
 (\protect\hyperlink{ref-Bengio2004}{2004})).
@@
 
+In machine learning tasks, selection of \(k\) for \(k\)-fold cross
+validation is largely an arbitrary decision between, e.g., 5 and 10.
+With the results of the following simulation, we make a less arbitrary
+recommendation of \(k\) with the aim of improving model selection
+accuracy.
+
 Consider a population of size N = 500,000 with five of \(P = 100\)
@@
 its mean squared error (\(\mathrm{MSE}\)) is
-\[\theta = \mathrm{MSE}[\mathrm{E}(Y)] = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.\]
+\[\theta = \mathrm{MSE}(f) = \frac{\sum_{i=1}^{N}[Y_i - \mathrm{E}(Y_i)]^2}{N}.\]
 
@@
 calculations of \(\mathrm{MSE}(\hat{\theta})\) such that a competing
-model \(\hat{f}\) has a lower \(\mathrm{\widehat{PE}}\) than the true
-model \(f\). This false model may have the lowest
+model \(\hat{f} \in F\) has a lower \(\mathrm{\widehat{PE}}\) than the
+true model \(f\). This false model may have the lowest
 \(\mathrm{\widehat{PE}}\) but its generalization error will be higher in
@@
 To estimate \(k_\mathrm{optimal}*\) for each \(n\), we first fit to the
-dataset \((k,A)\) for each sample size \(n\) a non-linear least squares
-model of the form \[\hat{A}(k) = Bk + C + Dk^{-1} + Ek^{-2}\] with
-constants \(B\) to \(E\) specific to each \(n\) and constants \(B\) and
-\(E\) set to zero for linear regression. We then draw a line \(L\)
+dataset \((k,A)\) for each sample size \(n\) a model of the form
+\[\hat{A}(k) = Bk + C + \frac{D}{k}\] with constants \(B\) to \(D\)
+specific to each \(n\). Certain more complicated models (e.g., with
+higher-order terms) may provide better approximation of \(A\) but make
+optimization by the following scheme non-trivial. We draw a line \(L\)
 through \((k_1, \hat{A}_1)\) and \((k_n, \hat{A}_n)\) and maximize the
 perpendicular distance \(d\) of each point on \(\hat{A}\) from \(L\), so
-that \[k_\mathrm{optimal}^* = {\arg\max}_k{d[\hat{A}(k)-L]}.\]
+that \begin{align*}
+k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[\hat{A}(k),L]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=m(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=\frac{(Bn+C+\frac{D}{n})-(2B+C+\frac{D}{2})}{n-2}(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},y-(Bn+C+\frac{D}{n})=(B-\frac{D}{2n})(x-n)]} \\
+                       &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]}. \\
+\end{align*} It is known that the distance \(d\) between a point
+\((k, \hat{A}(k))\) and a line of the form \[ax + by + c = 0\] is
+\begin{align*}
+d = \frac{|ak + b\hat{A}(k) + c|}{\sqrt{a^2+b^2}}
+\end{align*} so \begin{align*}
+k^*_{\mathrm{optimal}} &= {\arg\max}_k{d[Bk+C+\frac{D}{k},(B-\frac{D}{2n})x-y+C+\frac{D}{2}+\frac{D}{n}=0]} \\
+                       &= {\arg\max}_k{\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}}.
+\end{align*} Maximizing \(d\), knowing that \(n\) and \(k\) are always
+positive, we find that \begin{align*}
+\frac{d}{dk}d[\hat{A}(k),L] &= \frac{d}{dk}\left(\frac{|(B-\frac{D}{2n})k-(Bk+C+\frac{D}{k})+(C+\frac{D}{2}+\frac{D}{n})|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}}\right) \\
+                            &=\frac{|n||k|(2nD-Dk^2)(-Dk^2-2nD+nDk+2Dk)}{nk^3|-Dk^2-2nD+nDk+2Dk|\sqrt{D^2+4n^2}} = 0 \\
+                            &\Rightarrow k = 0, 2, \sqrt{2n}, n.
+\end{align*} Evaluating \(d\) at these four values of \(k\), we find
+that \begin{align*}
+d(0) &= \mathrm{DNE} \\
+d(2) &= 0 \\
+d(\sqrt{2n}) &=  \frac{|nD+2D-\sqrt{2n}D-D|}{\sqrt{(B-\frac{D}{2n})^2+(-1)^2}} \\
+d(n) &= 0.
+\end{align*} Since \(d(\sqrt{2n}) > 0\),
+\[k^*_{\mathrm{optimal}} = \sqrt{2n}\] which fits the relationship of
+\(k^*_{\mathrm{optimal}}\) vs.~\(n\) (Figure 9).
 
@@
 
-\caption{Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-7}
+\caption{Count of true linear regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k.}\label{fig:unnamed-chunk-7}
 \end{figure}
@@
 
-\caption{Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes.}\label{fig:unnamed-chunk-8}
+\caption{Count of true LASSO regression model selection by the lowest-error criterion vs. fold number for various sample sizes. Red line indicates optimal value of k.}\label{fig:unnamed-chunk-8}
 \end{figure}
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Bates2022}{}}%
+Bates, Stephen, Trevor Hastie, and Robert Tibshirani. 2022.
+{``Cross-Validation: What Does It Estimate and How Well Does It Do
+It?''} \url{https://arxiv.org/abs/2104.00673}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Bengio2004}{}}%

### 2023-07-20 plots/plots_lasso.R

@@ -33,15 +33,15 @@ variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 # loop settings
 n <- c(250, 500, 750, 1000)
 nsim <- 1000
-k <- list(n250 = c(2, seq(10, 100, 10), n[1]),
-          n500 = c(2, seq(10, 100, 10), n[2]),
-          n750 = c(2, seq(10, 100, 10), n[3]),
-          n1000 = c(2, seq(10, 100, 10), n[4])) # for n = 1000
+k <- list(n250 = c(2, seq(10, n[1], 10)),
+          n500 = c(2, seq(10, n[2], 10)),
+          n750 = c(2, seq(10, n[3], 10)),
+          n1000 = c(2, seq(10, n[4], 10))) # for n = 1000
 n_mod <- length(variables)
-out <- list(n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_250.rds"),
-            n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_500.rds"),
-            n750 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_750.rds"),
-            n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_1000.rds"))
+out <- list(n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
+            n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
+            n750 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_750_n_750.rds"),
+            n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
 titles <- list("n = 250", "n = 500", "n = 750", "n = 1000")
 plotsout <- list(list(), list(), list(), list())
 figures <- list()
@@ -110,6 +110,7 @@ figure6 <- annotate_figure(figures[[3]],
 counts <- list()
 corrects <- c()
 optplotsout <- list()
+pred <- list()
 for(i in 1:length(n)){
   
   # read results
@@ -121,18 +122,14 @@ for(i in 1:length(n)){
[…]

### 2023-07-20 plots/plots_linear.R

@@ -128,6 +128,7 @@ figure3 <- annotate_figure(figures[[3]],
 counts <- list()
 corrects <- c()
 optplotsout <- list()
+pred <- list()
 library(tidyr)
 for(i in 1:length(n)){
   
@@ -140,16 +141,14 @@ for(i in 1:length(n)){
   corrects <- counts[[i]] <- colSums(results == 2)
   
   # fit
-  nls.fit <- nls(corrects ~ c + d*k[[i]]^-1, start = list(c = 660, d = -14.5))
-  .c <- coef(nls.fit)[1]
-  .d <- coef(nls.fit)[2]
-  eq <- function(x){.c + .d*x^-1}
-  pred <- eq(2:n[[i]])
+  lm.fit <- lm(corrects ~ k + I(1/k), data = data.frame(corrects = corrects, k = k[[i]]))
+  lm.pred <- predict(lm.fit, data.frame(k = 2:n[[i]]))
+  pred[[i]] <- as.numeric(lm.pred)
   
   # plot
   optplotsout[[i]] <- ggplot(NULL)+
     geom_point(data = data.frame(f = k[[i]], corrects), aes(x = f, y = corrects), size = 0.1) +
-    geom_line(data = data.frame(f = 2:n[[i]], pred), aes(x = f, y = pred), size = 0.4, color = "black") +
+    geom_line(data = data.frame(f = 2:n[[i]], lm.pred), aes(x = f, y = lm.pred), size = 0.4, color = "black") +
     ggtitle(paste(titles[i])) +
     labs(x = NULL, y = NULL) +
     theme(panel.grid.major=element_blank(),
@@ -157,7 +156,8 @@ for(i in 1:length(n)){
           panel.background=element_blank(),
           axis.line=element_line(colour="black", size=0.4),
           axis.text=element_text(size=10, color="black"),
-          plot.title=element_text(hjust=0.5,size=10, color = "black"))
+          plot.title=element_text(hjust=0.5,size=10, color = "black")) +
+    geom_vline(xintercept = sqrt(n[[i]]), color = "red", size = 0.4)
 }
 
 optfigures <- ggarrange(optplotsout[[1]] + rremove("ylab") + rremove("xlab"),
@@ -182,15 +182,22 @@ figure7 <- annotate_figure(optfigures,
                            bottom = text_grob("k", size = 10),
[…]

### 2023-07-20 simulations/simulation_lasso_parallel.R

@@ -27,7 +27,7 @@ variables <- list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
 # 2, 10, 20, 30, ..., 100, n/2, n
-n <- 750
+n <- 1000
 k <- c(2, seq(10, n, 10))
 nsim <- 1000
 n_mod <- length(variables)
@@ -68,7 +68,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_n_1000.rds")
+out <- readRDS("/Users/angelos/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -114,19 +114,33 @@ for(sim in 1:nsim){
 corrects <- colSums(results == 2)
 plot(k, corrects)
 
+# lm
+lm.fit <- lm(corrects ~ k + I(1/k))
+lm.pred <- predict(lm.fit, data.frame(k = 2:n))
+lines(2:n, lm.pred)
+shapiro.test(lm.fit$residuals)
+
+
+
+
+
+
+
+
 # nls
-nls.fit <- nls(corrects ~ a*k + b + c*k^-1 + d*k^-2, start = list(a = -1, b = 660, c = 14.5, d = 0))
+nls.fit <- nls(corrects ~ a*k + b + c/k, start = list(a = -1, b = 660, c = 14.5))
 .a <- coef(nls.fit)[1]
 .b <- coef(nls.fit)[2]
 .c <- coef(nls.fit)[3]
-.d <- coef(nls.fit)[4]
-eq <- function(x){.a*x + .b + .c*x^-1 + .d*x^-2}
+eq <- function(m){.a*m + .b + .c/m}
 pred <- eq(2:n)
 lines(pred, type = "l")
 
 # elbow point
 library("smerc")
 elbow_point(2:n, pred)$x
-samp.n <- c(250, 500, 750, 1000)
-opt.k <- c(17, 17, 8, 5)
-plot(samp.n, opt.k/samp.n)
+samp.n <- c(100, 250, 500, 750, 1000)
[…]

### 2023-07-20 simulations/simulation_linear_parallel.R

@@ -26,7 +26,7 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- 400
+n <- 100
 nsim <- 1000
 k <- c(2, seq(10, n, 10)) # c(2, 10, 100, 500, 1000)
 n_mod <- length(variables)
@@ -46,7 +46,7 @@ library(doParallel)
 registerDoParallel(makeCluster(detectCores() - 1))
 
 # loop
-set.seed(2022)
+set.seed(2025)
 out <- foreach(j = 1:nsim, .combine = "list") %dopar% {
   ind <- sample(1:N, n, replace = FALSE)
   for(l in 1:length(k)){
@@ -65,7 +65,7 @@ out <- foreach(j = 1:nsim, .combine = "list") %dopar% {
 }
 
 # read results
-out <- readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_400_n_00.rds")
+out <- readRDS("~/Downloads/parallel_simulation_linear_output_k_2_100_n_100_d.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim), dimnames = list(rep("k", length(k)), rep("mod", n_mod), rep("sim", nsim)))
 
 # biassq + var = mse
@@ -123,11 +123,8 @@ lines(pred, type = "l")
 library("smerc")
 elbow_point(2:n, pred)$x
 samp.n <- seq(100, 1000, 100)
-opt.k <- c(14/100, 20/200, 24/300, 28/400, 32/500, 35/600, 37/700, 40/800, 42/900, 45/1000) # departing from LOOCV
-plot(samp.n, opt.k)
-
-
-
+opt.k <- c(14, 20, 24, 28, 32, 35, 37, 40, 42, 45) # departing from LOOCV
+plot(samp.n, opt.k/samp.n)
 
 
 

### 2023-09-02 manuscript/manuscript.Rmd
“Sept2”

@@
 urlcolor: cyan
-bibliography: /Users/angelos/Documents/research/optimal_k/manuscript/optimal_k_references.bib
+bibliography: /Users/angelos/optimal_k_git/manuscript/optimal_k_references.bib
 header-includes: \usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t} \usepackage[labelsep=period]{caption}
@@
 
+Note that the case of linear regression with $n = 100$ and number of features $p = 100$ is not displayed because there are $n - p - 1 = -1$ degrees of freedom, which is not a sufficient number for regression.
+The case of $n = 100$ and $p = 95$ degrees of freedom is similarly problematic and is not displayed because the high level of error in bias, variance, and $\mathrm{MSE}$ estimation associated with $n - p - 1 = 5$ degrees of freedom obscures important patterns in the rest of the data.
+
 As previous authors have noted (@Zhang2015), it is important to distinguish between possible goals of cross-validation that previously have been conflated (@Kohavi2001): estimation of $\mathrm{PE}$, in which case $k_\mathrm{optimal} = {\arg\min}_{k}(\mathrm{PE} - \mathrm{\widehat{PE}})$, and model selection, in which case $$k_\mathrm{optimal} = {\arg\max}_k(A)$$, which is related to the definition of $k_\mathrm{optimal}^*$ used in this study.

### 2023-09-02 manuscript/manuscript.tex
“Sept2”

@@
 
+Note that the case of linear regression with \(n = 100\) and number of
+features \(p = 100\) is not displayed because there are
+\(n - p - 1 = -1\) degrees of freedom, which is not a sufficient number
+for regression. The case of \(n = 100\) and \(p = 95\) degrees of
+freedom is similarly problematic and is not displayed because the high
+level of error in bias, variance, and \(\mathrm{MSE}\) estimation
+associated with \(n - p - 1 = 5\) degrees of freedom obscures important
+patterns in the rest of the data.
+
 As previous authors have noted (Zhang and Yang

### 2023-09-02 plots/plots_lasso.R

@@ -31,19 +31,21 @@ theta <- mean((y - Ey)^2)
 variables<-list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
-n <- c(250, 500, 750, 1000)
+n <- c(100, 250, 500, 750, 1000)
 nsim <- 1000
-k <- list(n250 = c(2, seq(10, n[1], 10)),
-          n500 = c(2, seq(10, n[2], 10)),
-          n750 = c(2, seq(10, n[3], 10)),
-          n1000 = c(2, seq(10, n[4], 10))) # for n = 1000
+k <- list(n100 = seq(2, 100, 2),
+          n250 = c(2, seq(10, n[2], 10)),
+          n500 = c(2, seq(10, n[3], 10)),
+          n750 = c(2, seq(10, n[4], 10)),
+          n1000 = c(2, seq(10, n[5], 10))) # for n = 1000
 n_mod <- length(variables)
-out <- list(n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
+out <- list(n100 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_100.rds"),
+            n250 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_250_n_250.rds"),
             n500 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_500_n_500.rds"),
             n750 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_750_n_750.rds"),
             n1000 = readRDS("~/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds"))
-titles <- list("n = 250", "n = 500", "n = 750", "n = 1000")
-plotsout <- list(list(), list(), list(), list())
+titles <- list("n = 100", "n = 250", "n = 500", "n = 750", "n = 1000")
+plotsout <- list(list(), list(), list(), list(), list())
 figures <- list()
 
 for(i in 1:length(out)){
@@ -86,8 +88,9 @@ for(i in 1:length(metrics)){
                             plotsout[[i]][[2]] + rremove("ylab") + rremove("xlab"),
                             plotsout[[i]][[3]] + rremove("ylab") + rremove("xlab"),
[…]

### 2023-09-02 plots/plots_linear.R

@@ -44,7 +44,7 @@ k <- list(n100 = c(2, seq(10, 100, 10)),
           n900 = c(2, seq(10, 900, 10)),
           n1000 = c(2, seq(10, 1000, 10)))
 n_mod <- length(variables)
-out <- list(n100 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_100_n_100.rds"),
+out <- list(n100 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_100_n_100_corrected.rds"),
             n200 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_200_n_200.rds"),
             n300 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_300_n_300.rds"),
             n400 = readRDS("~/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_400_n_400.rds"),
@@ -192,8 +192,8 @@ samp.n.linear <- seq(100, 1000, 100)
 opt.k.linear <- c(14, 20, 25, 28, 32, 35, 37, 40, 42, 45)
 opt.k.prop.linear <- opt.k.linear/samp.n.linear
 
-samp.n.lasso <- c(250, 500, 750, 1000)
-opt.k.lasso <- c(22, 32, 39, 45)
+samp.n.lasso <- c(100, 250, 500, 750, 1000)
+opt.k.lasso <- c(14, 22, 32, 39, 45)
 opt.k.prop.lasso <- opt.k.lasso/samp.n.lasso
 
 opt.k.raw <- ggplot(data.frame(f = samp.n.linear, opt.k.linear), aes(x = f, y = opt.k.linear))+

### 2023-09-02 simulations/simulation_lasso_parallel.R

@@ -27,8 +27,8 @@ variables <- list(1:3, 1:5, 1:20, 1:100, 6:100)
 
 # loop settings
 # 2, 10, 20, 30, ..., 100, n/2, n
-n <- 1000
-k <- c(2, seq(10, n, 10))
+n <- 100
+k <- seq(2, 100, 2)
 nsim <- 1000
 n_mod <- length(variables)
 
@@ -68,7 +68,7 @@ out <- foreach(j = 1:nsim, .combine = "list", .packages = "glmnet") %dopar% {
 }
 
 # read results
-out <- readRDS("/Users/angelos/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_1000_n_1000.rds")
+out <- readRDS("/Users/angelos/optimal_k_git/results/lasso/parallel_simulation_lasso_output_k_2_100_n_100.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim))
 
 # biassq + var = mse
@@ -119,10 +119,6 @@ lm.fit <- lm(corrects ~ k + I(1/k))
 lm.pred <- predict(lm.fit, data.frame(k = 2:n))
 lines(2:n, lm.pred)
 shapiro.test(lm.fit$residuals)
-
-
-
-
 
 
 

### 2023-09-02 simulations/simulation_linear_parallel.R

@@ -65,8 +65,11 @@ out <- foreach(j = 1:nsim, .combine = "list") %dopar% {
 }
 
 # read results
-out <- readRDS("~/Downloads/parallel_simulation_linear_output_k_2_100_n_100_d.rds")
+out <- readRDS("/Users/angelos/optimal_k_git/results/linear/parallel_simulation_linear_output_k_2_100_n_100.rds")
 theta.hat <- mse.hat <- array(unlist(out), dim = c(length(k), n_mod, nsim), dimnames = list(rep("k", length(k)), rep("mod", n_mod), rep("sim", nsim)))
+
+theta.hat[,c(4,5),] <- NA
+saveRDS(theta.hat, "parallel_simulation_linear_output_k_2_100_n_100_corrected.rds")
 
 # biassq + var = mse
 biassq <- (apply(theta.hat, c(1, 2), mean) - theta)^2
@@ -77,7 +80,7 @@ mse <- apply((theta.hat - theta)^2, c(1, 2), mean)
 library(plotrix)
 
 # bias^2 plots
-plot(k, biassq[, 1], type = "b", ylim = c(min(biassq), max(biassq)), ylab = "bias^2")
+plot(k, biassq[, 5], type = "b", ylab = "bias^2")
 points(k, biassq[, 2], type = "b", col = "blue")
 points(k, biassq[, 3], type = "b", col = "red")
 points(k, biassq[, 4], type = "b", col = "purple")
